# Data Sources {#ch:data-sources}

```{r, echo=TRUE, message=FALSE, results='hide', prompt=TRUE, size='tiny'}
  library(RCurl)
  library(data.table)
  library(fasttime)
  library(fImport)
  library(timeSeries)
  library(empfin)
  library(stringr)
  library(lubridate)
  library(quantmod)
  data(LiborRates)
```

This first chapter describes the data sources and the processing that
has been performed to create the data sets found in the package. In the
following chapters, we will use data from three asset classes:
commodities, equities and fixed income.

## Commodities Data

The CME group (www.cmegroup.com) provides a large quantity of data on
the products listed on its exchanges. Commodity prices are particularly
interesting in empirical finance because they exhibit high volatility,
seasonality, and many other specific features.

The following function downloads settlement data on commodity options
and futures traded on the NYMEX.

```{r, eval=FALSE, echo=TRUE}
  getNymexData <- function(type) {
    
  url <- 'ftp://ftp.cmegroup.com/pub/settle/'
  filename <- paste('nymex_', type, '.csv', sep='')
  fn <- paste(url, filename, sep='')

  curl <- getCurlHandle(ftp.use.epsv=FALSE)
  x <-  getURL(fn, curl=curl)

  #csv <- strsplit(x, '\n')[[1]]

  con <- textConnection(csv)
  tmp <- data.frame(read.csv(con, header=TRUE))
  close(con)

  if(type == 'future') {
    goodColumns <- c('PRODUCT.SYMBOL',
                   'CONTRACT.MONTH',
                   'CONTRACT.YEAR',
                   'CONTRACT',
                   'PRODUCT.DESCRIPTION',
                   'SETTLE',
                   'EST..VOL',
                   'TRADEDATE')

    goodNames <- c('Symbol', 'Month', 'Year', 'Contract',
                   'Description', 'Settle', 'Volume', 'TradeDate')
    
  }

  if(type == 'option') {
    goodColumns <- c('PRODUCT.SYMBOL',
                   'CONTRACT.MONTH',
                   'CONTRACT.YEAR',
                   'PUT.CALL',
                   'STRIKE',
                   'SETTLE',
                   'EST..VOL',
                   'TRADEDATE')

    goodNames <- c('Symbol', 'Month', 'Year',
                   'CP', 'Strike',
                   'Settle', 'Volume', 'TradeDate')
    }
  tmp <- tmp[,goodColumns]
  colnames(tmp) <- goodNames 
  tmp
  }
```

As an illustration, the following script downloads the NYMEX futures
data and extracts the Crude Oil (WTI) futures settlement prices. Running
the code on January 9, 2019, yields the following result:

```{r, eval=FALSE, echo=TRUE}
nymex.futures <- getNymexData('future')
tmp <- nymex.futures[nymex.futures\$Symbol == 'CL',]
head(tmp[,c('Month', 'Year', 'Contract', 'Settle')])

        Month Year Contract Settle
  4645     2 2019    CLG19  49.78
  4646     3 2019    CLH19  50.11
  4647     4 2019    CLJ19  50.49
  4648     5 2019    CLK19  50.93
  4649     6 2019    CLM19  51.36
  4650     7 2019    CLN19  51.74
```

For the record, the same function executed on October 11, 2012, yielded
the following data set:

```{r, eval=FALSE, echo=TRUE}
head(tmp[,c('Month', 'Year', 'Contract', 'Settle')])

  Month Year Contract Settle
  1998    11 2012    CLX12  91.25
  1999    12 2012    CLZ12  91.64
  2000     1 2013    CLF13  92.11
  2001     2 2013    CLG13  92.58
  2002     3 2013    CLH13  93.00
  2003     4 2013    CLJ13  93.30
```

The U.S. Energy Information Administration (www.eia.gov) publishes
historical data on spot and future prices. The data set in package was
obtained from that source.

## Libor Rates

The package also contains a data set of USD libor and swap rates
obtained from the US Federal Reserve archive (www.federalreserve.gov),
but this publication was discontinued at the end of October 2016.

The time series can be downloaded from a web browser, but can also be
downloaded programmatically. The site provides directions on how to
construct the URL corresponding to each particular data set.

In the example below, the URL is specific to the H15 table, with all
available deposit and swap rates included.

```{r, eval=FALSE, echo=TRUE}
  get.frb.url <- function(dtStart, dtEnd) {

      # Federal Reserve Board URL
      # Construct this URL at 'http://www.federalreserve.gov/datadownload
      
      url.1 <- paste(
      'https://www.federalreserve.gov/datadownload/Output.aspx?',
      'rel=H15&series=0b30e7214e1df1ba738a199d707702f1',
      '&lastobs=', sep='')
      url.2 <- '&filetype=csv&label=include&layout=seriescolumn'
      url.2 <- '&filetype=csv&label=include&layout=seriescolumn'
      url <- paste(url.1, 
      '&from=', as.character(dtStart, format='%m/%d/%Y'),
      '&to=',as.character(dtEnd, format='%m/%d/%Y'), 
      url.2, sep='')
      url
      }
```

The data is downloaded from the FRB web site and converted into a
object, with more explicit names. The resulting time series has daily
observations and 11 columns, with deposit rates of maturities 1, 3, and
6 months, and swap rates with maturities ranging from 1 to 30 years.

```{r, eval=FALSE, echo=TRUE}
  getLiborRates <- function(dtStart, dtEnd) {

    columns.dic = hash(keys=c("RIFLDIY01_N.B",
                            "RIFLDIY02_N.B",
                            "RIFLDIY03_N.B",
                            "RIFLDIY04_N.B",
                            "RIFLDIY05_N.B",
                            "RIFLDIY07_N.B",
                            "RIFLDIY10_N.B",
                            "RIFLDIY30_N.B",
                            "RILSPDEPM01_N.B",
                            "RILSPDEPM03_N.B",
                            "RILSPDEPM06_N.B"),
                       values = c('Swap1Y', 'Swap2Y',
                       'Swap3Y', 'Swap4Y', 'Swap5Y',
                       'Swap7Y', 'Swap10Y', 'Swap30Y',
                       'Libor1M', 'Libor3M', 'Libor6M'))

    # non-available data is marked 'ND' or 'NC' in the data set
    good.row <- function(x) length(grep('NC|ND', x)) == 0

    url <- get.frb.url(dtStart, dtEnd)
    
    curl <- getCurlHandle(ftp.use.epsv=FALSE)
    x <-  getURL(url, curl=curl)
    
    csv <- strsplit(x, '\n')[[1]]

    # skip 5 header rows at start of file
    con <- textConnection(csv[6:length(csv)])
    # colClasses='character' needed to avoid conversion to factors
    csv <- read.csv(con, colClasses='character', header=TRUE)
    close(con)

    ncol <- dim(csv)[2]
    indx <- apply(csv, 1, good.row)
    
    dt <- as.Date(csv\$Time.Period[indx])
    vx <- data.frame(csv[indx,2:ncol])
    for(i in seq(ncol-1)) vx[,i] = as.numeric(vx[,i])

    colNames = unlist(sapply(names(vx), function(x) columns.dic[[x]]),
                      use.names=FALSE)
    timeSeries(vx, dt, format='%Y-%m-%d', zone='GMT', FinCenter='GMT', 
    units=colNames)
  }
```

Executing this function on January 9, 2019 yields the following result.

```{r, eval=FALSE, echo=TRUE}
  > dtStart <- dmy('01Jan2010')
  > dtEnd <- dmy('01Jan2011')
  > ts = getLiborRates(dtStart, dtEnd)
  > head(ts[,c('Swap1Y', 'Swap10Y', 'Swap30Y')])
  GMT
             Swap1Y Swap10Y Swap30Y
  2010-01-04   0.65    3.93    4.51
  2010-01-05   0.61    3.87    4.46
  2010-01-06   0.60    3.92    4.52
  2010-01-07   0.57    3.92    4.53
  2010-01-08   0.55    3.94    4.58
  2010-01-11   0.53    3.91    4.56
```

The dataset in package has been constructed with this function, and
provides daily rates from July 2000 to October 2016.

As of this writing, current and historical Libor rates are available
from the Wall Street Journal Market Data Center (www.wsj.com). Swap
rates are available, for example, from Skandinaviska Enskilda Banken
(www.sebgroup.com).

## Sovereign Yield Curves

The European Central Bank publishes fitted curves of euro area AAA
sovereign yield. The data can be downloaded as a .csv file from this
institution’s statistical data warehouse (sdw.ecb.europa.eu). The
following script read, parse and display the par AAA euro government
yield curve for a set of dates.

```{r, read-ecb, cache=TRUE}
  # read, keep only par yield data
  ecb.file <- fread('./data/ecb-data.csv',
                    select=c('DATA_TYPE_FM',
                             'TIME_PERIOD',
                             'OBS_VALUE'))[substr(DATA_TYPE_FM,1,2) == 'PY']

  ecb.file$TIME_PERIOD <- as.Date(ecb.file$TIME_PERIOD)

  # convert to rectangular table, one column per maturity
  tmp <- dcast(ecb.file, TIME_PERIOD ~ DATA_TYPE_FM, value.var="OBS_VALUE")
  ecb.series <- tmp[,-1]
  # row names are characters
  row.names(ecb.series) <- tmp$TIME_PERIOD

  # decode maturity from column names
  tokens <- str_match(names(ecb.series), '^PY_(([0-9]+)Y)?(([0-9]+)M)?')

  # maturity expressed in months
  mat <- apply(tokens, 1, function(x) {
    tmp <- as.numeric(x[c(3,5)])
    tmp[is.na(tmp)] <- 0
    12*tmp[1] + tmp[2]
    })

  # rearrange columns by maturity
  sort.mat <- sort(mat, index.return=TRUE)
  setcolorder(ecb.series, neworder=sort.mat$ix)
```

```{r, label=plot-ecb, eval=FALSE, dependson='read-ecb'}
  nbObs <- dim(ecb.series)[1]
  indx <- seq(1, nbObs, 200)

  YC <- ecb.series[indx,]
  dtObs <- as.Date(row.names(ecb.series)[indx])
  nbSample <- length(indx)

  dtMat = dtObs[1] + months(sort.mat$x)
  plot(dtMat, YC[1,],type='l', ylim=c(-1,7), 
       xlim=c(dtObs[1], dtObs[nbSample]+months(last(sort.mat$x))),
       col=1, lwd=2, ylab='Euro AAA Govt Yield', xlab='Maturity date')
  abline(h=0, lty=2, lwd=2)
  for(i in seq(2,nbSample)) {
    dtMat = dtObs[i] + months(sort.mat$x)
    lines(dtMat, YC[i,],type='l', col=i, lwd=2)
  }
```

```{r, label=fig-plot-ecb,fig=T, echo=F}
  <<plot-ecb>>
```

## Equities Data

Time series of stock prices are readily available from free on-line
services. The provides a simple function to fetch
Open/High/Low/Close/Volume series from Yahoo Finance as an xts time
series:

```{r, load-and-plot, size='tiny',message=FALSE}
  getSymbols("F", src="yahoo")
```

and a good-looking bar chart is obtained by invoking

```{r, label=barchart-ford, eval=FALSE}
  barChart(F)
```

```{r, label=fig-bc-ford,fig=T, echo=FALSE}
  <<barchart-ford>>
```

For illustrative purpose, the package provides one time series of
daily prices for three stocks: IBM, Coca-Cola (KO) and Trans-Ocean
(RIG), and one Exchange Traded Fund: the Nasdaq 100 Trust (QQQQ).
