# The Simulation Framework {#sim-framework}

```{r setup, echo=FALSE, include=FALSE, cache=FALSE}
knitr::opts_chunk$set(
  cache = TRUE,
  dev.args = list(bg = "transparent"),
  tidy=TRUE,
  tidy.opts=list(width.cutoff=50)
)
source("formatters.R", local=knitr::knit_global())
```

```{r, echo=TRUE,results='hide', message=FALSE}
library(fInstrument)
library(empfin)
library(DynamicSimulation)
library(kableExtra)
```

```{r tufte::newthought("The")}``` simulation framework has two main components:

1.  A simulator, which can generate paths for the variables of interest
    (price of a asset, implied volatility, term structure of interest
    rate, etc.), according to a model that relates the variables to
    stochastic risk factors.

2.  A framework for expressing trading strategies, and in particular
    dynamic hedging policies.

This is again best explained by an example, and we describe next the use
of this framework for running a delta-hedging experiment. The main
elements of the design are discussed afterwards.

## Scenario Simulator

Assume that you want to simulate 500 price paths over a period of one
year, with 100 time steps. The price process will be log-normal with
annual volatility of $30\%$, starting from an initial value of $\$100$.

```{r, label=sim-1, echo=T, warning=FALSE}
  dtStart <- myDate('01jan2010')
  dtEnd <- myDate('01jan2011')
  nbSteps <- 100; nbPaths <- 500
```

Next, define the sequence of simulation dates and the volatility of the simulated log-normal process:

```{r, echo=T, warning=FALSE}
  dtSim <- seq(dtStart, dtEnd, length.out=nbSteps+1)
  sigma <- .3
```

Use a sobol sequence as random number generator, with antithetic variates, standardized to unit variance:

```{r, echo=T, warning=FALSE}
  tSpot <- pathSimulator(dtSim = dtSim, nbPaths=nbPaths,
      innovations.gen=sobolInnovations, path.gen=logNormal,
      path.param = list(mu=0, sigma=sigma), S0=100, antithetic = FALSE,
      standardization = TRUE, trace = FALSE)
  print(head(tSpot[,1:2]))
```

The output of the path simulator is a `r to_index("timeSeries", "classes")`. A plot of the first few paths is
displayed in figure \@ref(fig:price-plot). The function can generate simulated
values acording to various statistical processes; this is documented in
the vignette of the package.

```{r, label=price-plot, fig.cap="Simulated price paths under a log-normal diffusion process", fig.margin=TRUE, fig.height=4}
  plot(tSpot[,1:50], plot.type='single', ylab='Price', format="%b %d")
```

## A Delta Hedging Experiment

Having generated some scenarios for the stock price, let’s now simulate
the dynamic hedging of a European call option written on this stock,
using the Black-Scholes pricing model, with the implied volatility and
interest rate held constant.

Fist, we define the instrument to be hedged:

```{r, label=delta-hedge-0, echo=T}
  dtExpiry <- dtEnd

  underlying <- 'IBM'; K<-100

  a <- fInstrumentFactory("vanilla", quantity=1,
                    params=list(cp='c', strike=K,
                      dtExpiry=dtExpiry,
                      underlying=underlying,
                      discountRef='USD.LIBOR', trace=FALSE))
```

Next, we define the market data that will be held constant during the
simulation, and insert it in a `r to_index("DataProvider")`:

```{r, delta-hedge-2, echo=T, tidy=TRUE}
  base.env <- DataProvider()
  setData(base.env, underlying, 'Price', dtStart, 100)
  setData(base.env, underlying, 'DivYield', dtStart, .02)
  setData(base.env, underlying, 'ATMVol', dtStart, sigma)
  setData(base.env, underlying, 'discountRef', dtStart, 'USD.LIBOR')
  setData(base.env, 'USD.LIBOR', 'Yield', dtStart, .02)
```

At this stage, we can price the asset as of the start date of the
simulation:

```{r, label=value, echo=T}
  p <- getValue(a, 'Price', dtStart, base.env)
```

which gives a value of $p = `r round(p,2) `$.

Next, define the simulation parameters: we want to simulate a dynamic
hedging policy over 500 paths, and 100 time steps per path:

We use a child `r to_index("DataProvider")` to store the simulated paths. Data not found in the child `r to_index("DataProvider")`
will be searched for (and found) in the parent `r to_index("base.env")`.

```{r, delta-hedge-3, echo=T}
  sce.env <- DataProvider(parent=base.env)
  setData(sce.env, underlying, 'Price', time(tSpot), as.matrix(tSpot))
```

We can now run the delta-hedge strategy along each path:

```{r, delta-hedge-4, echo=T}
  assets = list(a)
  res <- deltaHedge(assets, sce.env,
                    params=list(dtSim=time(tSpot),
                    transaction.cost=0),trace=FALSE)
```

The result is a data structure that contains the residual wealth
(hedging error) per scenario and time step. The distribution of hedging
error at expiry is shown in Figure \@ref(fig:delta-hedge-41).

```{r, label=delta-hedge-41, fig.height=5, fig.cap='Distribution of residual wealth at expiry: delta hedge of a 1 year call option', fig.margin=TRUE}
  hist(tail(res$wealth,1), 50, xlab="Residual wealth at expiry", main='')
```

To better illustrate the hedging policy, let’s run a toy example with
few time steps. The function `r to_index("deltaHedge", "functions")` produces a detailed log of the hedging
policy, which is presented in Table \@ref(tab:delta-hedge-few-samples). For each time step,
the table show:

-   The stock price,
-   the option delta,
-   the option value,
-   the value of the replicating portfolio and the short bond position in that portfolio.

```{r, echo=T}
  dtSim <- time(tSpot)[seq(1, dim(tSpot)[1], 10)]
  res <- deltaHedge(assets, sce.env,
                    params=list(dtSim=dtSim,
                    transaction.cost=0),trace=FALSE)
  sim.table <- makeTable(1, res)
```

```{r, label=delta-hedge-few-samples, echo=FALSE}
  kable(sim.table, "latex", booktabs=T,
        digits=c(0,2,3,2,2,2),
        caption="Simulated value of a call option and its hedge portfolio over time")
```

### Design Considerations

Two design features are worth mentioning.

The generation of the scenarios is independent from the expression of
the dynamic trading strategies. Remember that every data element stored
in a `r to_index("DataProvider")` is a `r to_index("timeSeries")`. Since all the calculations on `r to_index("fInstrument")`
are vectorized, there is no
difference between performing a calculation on a scalar, or performing a
simulation on multiple scenarios.

The second aspect is the use of parent/child relationships among
`r to_index("DataProvider")` objects. All the market data that is held constant in the simulation is
stored in the parent `r to_index("DataProvider", "classes")`. The data that changes from simulation to
simulation is stored in the child `r to_index("DataProvider", "classes")`, and this is the object that is
passed to the simulator. When a piece of data is requested from the
child `r to_index("DataProvider", "classes")`, the following logic is applied:

1.  First look for the data in the current `r to_index("DataProvider", "classes")` (the object passed as
    argument to the simulator)
2.  if not found, look for the data in the parent `r to_index("DataProvider", "classes")`.
3.  and so forth: the logic is recursive.

This behavior is inherited from the built-in `r to_index("environment")`.
