# Arbitrage-Free Pricing and Risk Neutral Valuation {#arb-free-pricing}

```{r, init-discreteModels,echo=TRUE,results='hide'}
  library(fBasics)
  library(xtable)
  library(empfin)
```

## Arbitrage-free Pricing

We are exclusively concerned about the pricing of redundant securities,
i.e. relative value pricing. Financial economics tries to explain the
pricing of underlying securities, mathematical finance is about the
relative value pricing of derivatives securities.

### An Introductory Example

Consider three produce baskets of apples and oranges.

   Basket   Apples   Oranges   Price (€)
  -------- -------- --------- -----------
     B1       2         3          4
     B2       3         2          5
     B3       2         2          ?


Is € 3.5 a fair price for basket B3? The answer is no: I can buy 5
baskets B3 and sell 2 baskets B1 and 2 baskets B1 for a riskless profit
of €0.5. The fair price for B3 is 3.6 €. We have determined the
arbitrage-free price for B3 by constructing a replication out of baskets
B1 and B2. This is the essence of derivatives pricing.

### Arrow-Debreu Securities

Imagine an economy that can evolve between time $t=0$ and $t=1$ to take
one out of 3 possible states. There is a consensus to attribute
probability $p_i$ to the occurence of state $i$. The interest rate is 0.

We next define 3 securities; each one pays a certain pattern of cashflow
according to the future state, as pictured in figure \[fig:3states\].

```{tikz, ThreeStates, fig.cap="Cashflows from 3 securities", 
fig.align="center"}
\usetikzlibrary{calc}
\usetikzlibrary{positioning}
\usetikzlibrary{arrows}
\begin{tikzpicture}

\node{$S_1$}[grow' =right,sibling distance=1cm] child {node {0}} child {node
{3}} child {node {1}};

\node{$S_2$}[grow'=right,sibling distance=1cm] child {node {1}} child {node
{2}} child {node {1}};

\node{$S_3$}[grow'=right,sibling distance=1cm] child {node {1}} child {node
{0}} child {node {2}};

\end{tikzpicture}
```
Assume that prices at time $t=0$ are: $S_1 = 1.3$, $S_2 = 1.25$,
$S_3 = 1.3$. The price of these securities is determined by several
factors:

1.  the preference of market participants for earning a payoff in one
    state versus another: in a risk-adverse economy, earning 1 euro when
    the aggregate wealth is low is more valuable than earning the same
    amount when the aggregate wealth is high

2.  the preference for holding money today rather than at time $T$

3.  the likelihood of each state

We are now interested in the price of securities that pays 1 euro if
state $i$ is realized, and nothing otherwise. Such securities are called
Arrow-Debreu securities, and can be represented by a vector giving the
value of such security in each future state of the world.

The value of the first Arrow-Debreu security is determined by solving
the following system:

$$\begin{pmatrix}
    0 & 1 & 1 \\
    3 & 2 & 0 \\
    1 & 1 & 2
\end{pmatrix} W_1 =
\begin{pmatrix}
1 \\
0 \\
0
\end{pmatrix}$$


Replication of the first Arrow-Debreu security $$\begin{pmatrix}
    0 & 1 & 1 \\
    3 & 2 & 0 \\
    1 & 1 & 2
\end{pmatrix} W_1 =
\begin{pmatrix}
1 \\
0 \\
0
\end{pmatrix}$$

which yields

```{r, label=AD-Price, echo=T }
  P <- matrix(c(0,3,1,1,2,1,1,0,2),3,3)
  A1 <- matrix(c(1,0,0),3,1)
  W1 <-  inv(P) %*%  A1
  W1
```

The value of the first Arrow-Debreu security is thus:

```{r, label=S1-Price}
  S <- matrix(c(1.3, 1.25, 1.3),3,1)
  a1 <- t(W1) %*% S
  print(a1[1,1])
```

Similarly, the prices $a_2$ and $a_3$ of the other two Arrow-Debreu
securities are computed by:

```{r, label=AD-Price2}
  W2 <-  inv(P) %*% matrix(c(0,1,0),3,1)
  a2 <- t(W2) %*% S
  W3 <-  inv(P) %*% matrix(c(0,0,1),3,1)
  a3 <- t(W3) %*% S
```

The replicating portfolios and prices of the 3 A-D assets are summarized
in table \[tab:state-rep\].

```{r, results='asis',echo=FALSE,fig=FALSE}
  z <- as.matrix(cbind(W1, W2, W3, rbind(a1, a2, a3)))
  colnames(z) <- c('W1', 'W2', 'W3', 'Price')
  xres<-xtable(z, digits=2)
```

```{r, results='asis',echo=FALSE,fig=FALSE}
  print.xb(xres, include.colnames=T)
```

\[tab:state-rep\]


```{r, results='asis',echo=FALSE,fig=FALSE}
  print.xb(xres, include.colnames=T)
```

A portfolio made of the three Arrow-Debreu securities guarantees a
payoff of 1 euro. Therefore, this collection can be priced at time $t=0$
by discounting the payoff at the risk-free rate (0 for now). Thus, we
must have:

$$\sum a_i = 1$$

Which is indeed the case:

```{r, echo=T}
  print(a1+a2+a3)
```

### The Price of Traded Securities

The price of Arrow-Debreu securities is determined by the price of
traded securities. We now consider how these price are determined.

The current price of an asset depends on the future payoffs, and on the
states in which these payoffs occur: if the payoffs of an asset are
positively correlated with the aggregate market value, it will be worth
less, everything else being equal, than an asset whose payoffs are
negatively correlated with the aggregate market value. The capital asset
pricing model formalizes this observation. Let

$S_t$

:   Asset price at time $t$

$M_t$

:   Aggregate market wealth at time $t$

$R_s$

:   Return on asset: $S_T/S_0$

$R_m$

:   Market return: $M_T/M_0$

$R_f$

:   Risk-free return: $1+r$

The model relates the expected return of a security to the beta value:

$$E(R_s) = R_f + \beta [ E(R_m) - R_f ]$$

where

$$\beta = \frac{\mbox{Cov}(R_s, R_m)}{\mbox{Var}(R_m)}$$

In that framework, it can be shown that $S_0$, the current asset price,
is given by:

$$S_0 = \frac{E(S_T)-\lambda \mbox{Cov}(S_T, M_T)}{R_f}
    \label{eq:capm1}$$

where $\lambda$ is the market price of risk times the current market
value $M_0$:

$$\lambda = \frac{M_0 [ E(R_m)-R_f ]}{\mbox{Var}(M_T)}$$

In a complete market, the asset and the market portfolio can be
expressed in terms of Arrow-Debreu securities:

$$S_0 = \sum_i V_i a_i$$

$$M_0 = \sum_i U_i a_i$$

Assume that the states are ordered in order of increasing aggregate
wealth. We have:

$$\begin{aligned}
    E(S_T) & = & \sum_i V_i p_i \\
    E(M_T) & = & \sum_i U_i p_i \\
    E(S_T M_T) & = & \sum_i U_i V_i p_i\end{aligned}$$

Substitute in \[eq:capm1\] to get:

$$S_0 = \sum_i V_i d_i p_i$$

where the discount factor $d_i$ is given by:

$$d_i = \frac{1-\lambda(U_i - E(M_T))}{R_f}
    \label{eq:capm2}$$

Equation \[eq:capm2\] shows that as aggregate wealth $U_i$ increases,
the discount factor decreases. An asset is more valuable, everything
else being equal, if its payoffs occur in the states where $U_i$ is low,
and therefore where the discount factor is high.

To generalize: any factor that affects supply and demand for traded
securities, and the market price of risk, will have a direct influence
on the Arrow-Debreu prices and therefore on the risk neutral
probabilities.

As noted by Derman and Taleb([@Derman2005]),

> The Nobel committee upon granting the Bank of Sweden Prize in honour
> of Alfred Nobel, provided the following citation: “Black, Merton and
> Scholes made a vital contribution by showing that it is in fact not
> necessary to use any risk premium when valuing an option. This does
> not mean that the risk premium disappears; instead it is already
> included in the stock price.” It is for having removed the effect of
> $\mu$ \[the stock drift\] on the value of the option, and not for
> rendering the option a deterministic and riskless security, that their
> work is cited.

### Complete Market

A complete market is a market where all Arrow-Debreu securities can be
traded, and therefore any payoff profile can be replicated as a
portfolio of Arrow-Debreu securities. The existence of this replicating
portfolio imposes a unique arbitrage-free price for any payoff. Going
back to the elementary example above, consider now a new security that
has the payoff profile illustrated in figure \[fig:new-sec\].

\[grow’=right,sibling distance=1cm\] child [node [1]{}]{} child [node
[-0.5]{}]{} child [node [1]{}]{};


\[grow’=right,sibling distance=1cm\] child [node [1]{}]{} child [node
[-0.5]{}]{} child [node [1]{}]{};

This security is equivalent to a portfolio of three Arrow-Debreu
securities, and is worth

```{r, echo=T}
  V <- a1 - 0.5*a2 + a3
```

If its market price of $S_4$ is less than $V=`r round(V, 2) `$, you
can earn a riskless profit by buying a unit of $S_4$ and selling the
portfolio $P$.

In general, a security with payoff $X_i$ in state $i$ is worth:

$$\sum_i a_i X_i$$

with

$$\sum_i a_i = 1$$

and

$$a_i >0, \ \ \forall i$$

and we can interpret the Arrow-Debreu prices as probabilities. Since
preferences no longer play a role, we call them “risk-neutral”
probabilities.

### Equivalent Probability Measures

Two probability measures $p$ and $q$ are equivalent if they are
consistent with respect to possible and impossible outcome:

$$p_i >0 \Leftrightarrow q_i >0$$

Let $p$ be the real probability measure and $q$ be the risk-neutral
measure. It is easy to show that the two measures must be equivalent.

Consider a state $i$ such that $p_i = 0$. then the corresponding
Arrow-Debreu security cannot cost anything, or a riskless profit could
be gained by selling this security. Thus, $q_i=0$. A similar argument
applies for the case $p_i>0$.

### The Impact of Interest Rate

What happens to Arrow-Debreu prices and risk-neutral probabilities when
interest rate is not null? In the presence of interest rate, the value
of a complete set of Arrow-Debreu securities must be

$$\sum_i a_i = e^{-rT}$$

The risk-neutral probabilities are now defined as:

$$q_i = a_i e^{rT}$$

so that we still have $\sum_i q_i = 1$. As before, an arbitrary security
with payoff $X_i$ in state $i$ is worth,

$$P(X) = e^{-rT} \sum_i X_i q_i$$

or,

$$P(X) = e^{-rT} E^Q[X]$$

Risk-neutral probabilities are compounded Arrow-Debreu prices. Here, for
expository purpose, we have derived risk-neutral probabilities from
state prices, but in practice, we will do the opposite: to obtain state
prices from risk-neutral probabilities.

### Trading Strategy and Dynamic Completeness

Let’s now consider an economy where decisions can be made at various
stages. This economy is illustrated in Figure \[fig:bin-tree-0\]. At the
second time step, we have 3 distinct states, i.e. 3 Arrow-Debreu
securities. If we only had one time step, we would need 3 linearly
independent assets to construct these securities. Now, because of the
intermediate trading opportunity, we may be able to construct the
Arrow-Debreu securities with fewer (i.e. two) underlying assets. A
market in which every Arrow-Debreu security can be constructed with a
self-financing trading strategy is called dynamically complete.

### Discounted Asset Prices as Martingales

Let’s consider again a two-stage economy. What can we say at time 0
about the expected value of a derivative at a future time $t$?

$$E_0^Q[P_t(X)], \forall t \geq 0$$

$$E_0^Q[P_t(X)] = e^{-2r}E_0^Q[X]$$

Now let’s consider the expected price at $t=1$:

$$\begin{aligned}
E_0^Q[P_1(X)] & = & E_0^Q[e^{-r}E_1^Q[X]] \\
E_0^Q[P_1(X)] & = & E_0^Q[e^{-r}E_0^Q[X]] \\
E_0^Q[P_1(X)] & = & e^{-r} E_0^Q[X]\end{aligned}$$

Similarly,

$$\begin{aligned}
E_0^Q[P_2(X)] & = & E_0^Q[E_2^Q[X]] \\
E_0^Q[P_2(X)] & = & E_0^Q[X]\end{aligned}$$

The price of each state-dependent payoff grows at the risk-free rate,
simply because this growth rate is incorporated in the definition of
each risk-neutral probability with which we weight the state-dependent
payoffs.

The expected price of any asset (as seen from time $t=0$) grows at the
risk-free rate. Not the price, but the expectation of the price.
Generalizing, we have:

$$P_0(X) = e^{-rt}E_0[P_t(X)], \ \forall t>0$$
