# Stability of the Binomial Model

```{r setup, echo=FALSE, include=FALSE, cache=FALSE}
knitr::opts_chunk$set(
  cache = TRUE,
  dev.args = list(bg = "transparent"),
  tidy=TRUE,
  tidy.opts=list(width.cutoff=45)
)
```

```{r, label=discmodbinomial2-setup, echo=FALSE, message=FALSE, results='hide'}
  library(fOptions)
  library(fExoticOptions)
  library(fAsianOptions)
  library(empfin)
```

`r tufte::newthought("We")` would like to verify the convergence of the Cox-Ross model to the
Black-Scholes price as the number of steps $N$ increases. This can be
investigated with the following script:

```{r, label=Binomial-price, echo=TRUE}
  Strike <- 100; Spot <- 100; T1 <- 1/4 
  r <- .05; b <- .05; sigma <- .3

  NN <- seq(20, 100, 1)
  nb <- length(NN)
  res <- matrix(nrow=nb, ncol=2)
  bs <- rep(0,2)

  # The Black-Scholes price
  bs[1] <- GBSOption(TypeFlag='c', S=Spot, X=Strike, 
      Time=T1, r=r, b=b, sigma=sigma)@price
  bs[2] <- GBSOption(TypeFlag='c', S=Spot, X=Strike+10, 
      Time=T1, r=r, b=b, sigma=sigma)@price

  # Binomial price, function of number of steps
  res[, 1] <- sapply(NN, function(n) CRRBinomialTreeOption(
          TypeFlag='ce', S=Spot, X=Strike, Time=T1, 
          r=r, b=b, sigma=sigma, n)@price)

  res[, 2] <- sapply(NN, function(n) CRRBinomialTreeOption(
          TypeFlag='ce', S=Spot, X=Strike+10, Time=T1, 
          r=r, b=b, sigma=sigma, n)@price)
```

A plot of the prices as a function of the number of steps
(Figure \[fig:bin-conv\]) shows an oscillating pattern:

```{r, binomial-plot, fig=FALSE, echo=T, eval=FALSE}
  par(mfrow=c(1, 2))
  plot(NN, res[, 1], type='l',main='ATM', xlab='Number of steps', ylab='price', col='blue', ylim=c(min(res[,1]), max(res[,1])))
  abline(h=bs[1], lwd=2, col='red')

  plot(NN, res[, 2], type='l',main='OTM', xlab='Number of steps', ylab='price', col='blue', ylim=c(min(res[,2]), max(res[,2])))
  abline(h=bs[2], lwd=2, col='red')
  par(mfrow=c(1, 1))
```

```{r, label=fig-bin-conv-a,echo=FALSE, fig.width=6, fig.height=4}
  <<binomial-plot>>
```

Other binomial algorithms, such as Tian’s, exhibit a similar pattern, as
evidenced in Figure \[fig:bin-conv-2\]. The horizontal line marks the
Black-Scholes price.

```{r,tian-price, echo=FALSE}
  res[, 1] <- sapply(NN, function(n) TIANBinomialTreeOption(
          TypeFlag='ce', S=Spot, X=Strike, Time=T1, 
          r=r, b=b, sigma=sigma, n)@price)
  res[, 2] <- sapply(NN, function(n) TIANBinomialTreeOption(
          TypeFlag='ce', S=Spot, X=Strike+10, Time=T1, 
          r=r, b=b, sigma=sigma, n)@price)
```

```{r, binomial-plot2, fig=FALSE, echo=FALSE, eval=FALSE}
  par(mfrow=c(1, 2))
  plot(NN, res[, 1], type='l',main='ATM', xlab='Number of steps', ylab='price', col='blue', ylim=c(min(res[,1]), max(res[,1])))
  abline(h=bs, lwd=2, col='red')

  plot(NN, res[, 2], type='l',main='OTM', xlab='Number of steps', ylab='price', col='blue', ylim=c(min(res[,2]), max(res[,2])))
  abline(h=bs, lwd=2, col='red')
  par(mfrow=c(1, 1))
```

```{r, label=fig-bin-conv-2-a,echo=FALSE, fig.width=6, fig.height=4}
  <<binomial-plot2>>
```

This sawtooth pattern is due to the position of the strike relative to
the sequence of nodes at expiry; 
we describe below some computational strategies 
for smoothing these oscillations and speeding up the convergence of
binomial trees. See
@Joshi2007 for an extensive survey of binomial models with
improved convergence properties.

Since the oscillations are caused by the variations in the relative
position of the strike with respect to nodes at expiry, a natural
strategy, introduced by @Leisen1996, is to construct the tree such that
the strike coincides with a node. This is achieved by setting
$$\mu = \frac{1}{T} \log \left(\frac{K}{S_0} \right)$$

The resulting tree is centered on $K$ in log space. The pricing method
is implemented as follows:

```{r, label=CoxRossWithBS, echo=T}
  CRRWithDrift <- function (TypeFlag = c("ce", "pe", "ca", "pa"), S, X, Time, r, 
      mu, sigma, n) 
  {
      TypeFlag = TypeFlag[1]
      z = NA
      if (TypeFlag == "ce" || TypeFlag == "ca")
          z = +1
      if (TypeFlag == "pe" || TypeFlag == "pa") 
          z = -1
      if (is.na(z)) 
          stop("TypeFlag misspecified: ce|ca|pe|pa")
      dt = Time/n
      u = exp(mu*dt + sigma*sqrt(dt))
      d = exp(mu*dt - sigma*sqrt(dt))
      
      p = (exp(r * dt) - d)/(u - d)
      Df = exp(-r * dt)
     
      # underlying asset at step N-1
      ST <- S*(d^(n-1))*cumprod(c(1, rep((u/d), n-1)))
      # at step (n-1), value an European option of maturity dt
      BSTypeFlag <- substr(TypeFlag,1,1)
      OptionValue <- GBSOption(BSTypeFlag, ST, X, dt, r, b, sigma)@price

      if (TypeFlag == "ce" || TypeFlag == "pe") {
          for (j in seq(from = n - 2, to = 0, by = -1))
            OptionValue <- (p*OptionValue[2:(j+2)] + 
                            (1-p)*OptionValue[1:(j+1)])*Df
      }
      
      if (TypeFlag == "ca" || TypeFlag == "pa") {
          for (j in seq(from = n - 2, to = 0, by = -1)) {
            ContValue <- (p*OptionValue[2:(j+2)] + (1-p)*OptionValue[1:(j+1)])*Df
            ST <- S*(d^j)*cumprod(c(1, rep((u/d), j)))
            OptionValue <- sapply(1:(j+1), function(i) max(ST[i]-X, ContValue[i]))
          }
      }

      OptionValue[1]
  }
```

Convergence of the model as $N$ increases is significantly improved, as
evidenced by the graphs in Figure \[fig:CRRWithDrift\].

```{r, label=CRRWithDrift-plot-calc, echo=FALSE}
  mu = (1/T1) * log(Strike/Spot)
  res[, 1] <- sapply(NN, function(n) CRRWithDrift(
          TypeFlag='ce', S=Spot, X=Strike, Time=T1, 
          r=r, mu=mu, sigma=sigma, n))
  res[, 2] <- sapply(NN, function(n) CRRWithDrift(
          TypeFlag='ce', S=Spot, X=Strike+10, Time=T1, 
          r=r, mu=mu, sigma=sigma, n))
```

```{r, CRRWithDrift-plot, fig=FALSE, eval=FALSE, echo=FALSE}
  par(mfrow=c(1, 2))
  plot(NN, res[, 1], type='l',main='ATM', xlab='Number of steps', ylab='price', col='blue', ylim=c(min(c(bs[1], res[,1]))-.02, max(res[,1])))
  abline(h=bs[1], lwd=2, col='red')

  plot(NN, res[, 2], type='l',main='OTM', xlab='Number of steps', ylab='price', col='blue', ylim=c(min(c(bs[2], res[,2]))-.02, max(res[,2])))
  abline(h=bs[2], lwd=2, col='red')
  par(mfrow=c(1, 1))
```

```{r, label=fig-crrwithdrift,echo=FALSE, fig.width=6, fig.height=4}
  <<CRRWithDrift-plot>>
```

In the context of an American option, note that if the option has not
been exercised at step $N-1$, the option is now European, and can be
priced at these nodes with the Black-Scholes model, rather than with the
backward recursion from step $N$ (the expiry date). This simple
modification smooths the option value at step $N-1$ and cancels the
oscillations, as illustrated in figure \[fig:bin-conv-crr-bs\], but at
the price of a substantial increase in computation time.

```{r, label=CoxRossWithBS-2, echo=T}
  CRRWithBS <- function (TypeFlag = c("ce", "pe", "ca", "pa"), S, X, Time, r, 
      b, sigma, n) 
  {
      TypeFlag = TypeFlag[1]
      z = NA
      if (TypeFlag == "ce" || TypeFlag == "ca")
          z = +1
      if (TypeFlag == "pe" || TypeFlag == "pa") 
          z = -1
      if (is.na(z)) 
          stop("TypeFlag misspecified: ce|ca|pe|pa")
      dt = Time/n
      u = exp(sigma * sqrt(dt)); d = 1/u
      p = (exp(b * dt) - d)/(u - d)
      Df = exp(-r * dt)
     
      # underlying asset at step N-1
      ST <- S*(d^(n-1))*cumprod(c(1, rep((u/d), n-1)))
      # at step (n-1), value an European option of maturity dt
      BSTypeFlag <- substr(TypeFlag,1,1)
      OptionValue <- GBSOption(BSTypeFlag, ST, X, dt, r, b, sigma)@price

      if (TypeFlag == "ce" || TypeFlag == "pe") {
          for (j in seq(from = n - 2, to = 0, by = -1))
            OptionValue <- (p*OptionValue[2:(j+2)] + 
                            (1-p)*OptionValue[1:(j+1)])*Df
      }
      
      if (TypeFlag == "ca" || TypeFlag == "pa") {
          for (j in seq(from = n - 2, to = 0, by = -1)) {
            ContValue <- (p*OptionValue[2:(j+2)] + (1-p)*OptionValue[1:(j+1)])*Df
            ST <- S*(d^j)*cumprod(c(1, rep((u/d), j)))
            OptionValue <- sapply(1:(j+1), function(i) max(ST[i]-X, ContValue[i]))
          }
      }

      OptionValue[1]
  }
```

```{r, label=CRRWithBS-plot, eval=FALSE, echo=FALSE}
  res[,1] <- sapply(NN, function(n) CRRBinomialTreeOption(TypeFlag='ca', S=Spot, X=Strike, Time=T1, r=r, b=b, sigma=sigma, n)@price)
  res[,2] <- sapply(NN, function(n) CRRWithBS(TypeFlag='ca', S=Spot, X=Strike, Time=T1, r=r, b=b, sigma=sigma, n))

  plot(NN, res[,1], type='l', xlab='Number of steps', ylab='price', col='blue')
  abline(h=bs[1], lwd=2, col='red')
  lines(NN, y=res[,2], type='l', lwd=2, col='green')
  # legend('topright', c("Cox-Ross", "BS", "Cox-Ross+BS step"), lty=c(1,1), 
  #         col=c('blue', 'red', 'green'),bg="gray90")
```

```{r, label=fig-CRRWithBS,echo=FALSE, fig.width=5, fig.height=4}
<<CRRWithBS-plot>>
```
