# Price and Greeks in the Black-Scholes Model

```{r, label=load-BS-Study,echo=FALSE,results='hide'}
  library(fOptions)
  library(fExoticOptions)
  library(fAsianOptions)
  library(fInstrument)
  library(DynamicSimulation)
  library(empfin)
```

In this chapter, we consider the price and risk indicators of several
types of options. We are particularly interested in the higher order
risk indicators, such as the Gamma. We will see later that this
indicator has a major impact on our ability to dynamically hedge such
instruments.

Solving the Pricing Equation
----------------------------

A fairly rigorous derivation of the Black-Scholes equation can be
obtained without using the tools of stochastic calculus. Recall from the
previous section that under the risk-neutral probability, the discounted
value is a martingale:

$$S_0 = e^{-rT} E^Q(S_T)$$

where $S_T$ is a log-normal variable that follows the process:

$$S_T = S_0 e^{(r-\frac{\sigma^2}{2})T + \sigma \sqrt{T}Y}$$

with $Y ~ N(0,1)$. The value of any derivative is the discounted
expected payoff. For a call option, we have:

$$\begin{aligned}
    c(S, 0) &=& e^{-rT} E^Q [ \left( S_0 e^{(r-\frac{\sigma^2}{2})T + \sigma \sqrt{T}Y}-K \right)^+ ] \\
    &=& \frac{e^{-rT}}{\sqrt{2\pi}} \int_{-\infty}^{\infty} \left( S_0 e^{(r-\frac{\sigma^2}{2})T + \sigma \sqrt{T}Y}-K \right)^+ e^{-x^2/2} dx\end{aligned}$$

Now compute the bounds of integration:

$$\begin{aligned}
    & & S_0 e^{(r-\frac{\sigma^2}{2})T + \sigma \sqrt{T}x} \geq K \\
    & <=> & e^{\sigma \sqrt{T} x} \geq \frac{K}{S_0} e^{-(r-\frac{\sigma^2}{2})T} \\
    & <=> & x \geq \frac{1}{\sigma \sqrt{T}} \left( \ln(\frac{K}{S_0}) -(r-\frac{\sigma^2}{2})T \right)\end{aligned}$$

Let

$$T_1 = \frac{1}{\sigma \sqrt{T}} \left( \ln(\frac{K}{S_0}) -(r-\frac{\sigma^2}{2})T \right)$$

$$\begin{aligned}
    c(S, 0) &=&  \frac{e^{-rT}}{\sqrt{2\pi}} \int_{T_1}^{\infty} \left( S_0 e^{(r-\frac{\sigma^2}{2})T + \sigma \sqrt{T}x}-K \right) e^{-x^2/2} dx \\
    &=& \frac{e^{-rT}}{\sqrt{2\pi}} \int_{T_1}^{\infty} S_0 e^{(r-\frac{\sigma^2}{2})T + \sigma \sqrt{T}x} e^{-x^2/2} dx - K  \frac{e^{-rT}}{\sqrt{2\pi}} \int_{T_1}^{\infty}  e^{-x^2/2} dx \\
    &=& A - B\end{aligned}$$

$$B = K e^{-rT} (1-N(T_1))$$

Apply the change of variable $y = x-\sigma \sqrt{T}$ to get

$$\begin{aligned}
    A &=&  \frac{e^{-rT}}{\sqrt{2\pi}} \int_{T_1-\sigma \sqrt{T}}^{\infty} S_0 e^{-y^2/2} dy \\
      &=& S_0 (1-N(T_1 - \sigma \sqrt{T}))\end{aligned}$$

where $N(x)$ is the cumulative normal distribution. Using the identity
$N(x)+N(-x)=1$, one gets the Black-Scholes formula in its standard form.

$$C_{\mbox{BS}}(S,0) = S_0N(d_1) - Ke^{-rT}N(d_2) \,
\label{eq:bs-call}$$

with

$$\begin{aligned}
    d_1 &=& \frac{\ln(S/K) + (r + \sigma^2/2)T}{\sigma\sqrt{T}} \\
    d_2 &=& d_1 - \sigma\sqrt{T}.\end{aligned}$$

The value of the put can be obtained from the call-put parity
relationships.

In the following discussion about the management of market risk of
option portfolios, we use three risk indicators, also known as greeks:

-   Delta, which is the derivative of price with respect to the price of
    the underlying asset,

-   Gamma, which is the second derivative of price with respect to spot,
    and

-   Vega, which is the derivative of price with respect to the
    volatility.

See, for example Hull \[[@Hull1997]\] for a detailed discussion of these
risk indicators.

An Option Dashboard
-------------------

We start the chapter by constructing a “dashboard” which illustrates the
properties of various types of European options. These displays are
inspired by similar figures developped by D. Eddelbuettel
\[[@Eddelbuettela]\] for the RQuantLib project. The class provides a
generic way of writing such function.

The dashboard has four panels, which display the price, delta, gamma and
vega of the option under consideration. Each panel has a family of
curves varying in color from yellow to green to blue. This family of
curves pictures the evolution of the price and of the greeks for varying
levels of volatility. The yellow curves corresponding to a high
volatility conditions, and at the other end, the blue line corresponding
to low volatility conditions.

```{r, label=OptionDashBoard, echo=TRUE, results='hide'}
  OptionDashBoard <- function(inst, base.env, dtCalc, und.seq, vol.seq) {
    sce <- t(as.matrix(und.seq))
    underlying <- inst@params$underlying
    setData(base.env, underlying, 'Price', dtCalc, sce)

  p <- matrix(nrow=length(und.seq), ncol=length(vol.seq))
  d <- matrix(nrow=length(und.seq), ncol=length(vol.seq))
  g <- matrix(nrow=length(und.seq), ncol=length(vol.seq))
  v <- matrix(nrow=length(und.seq), ncol=length(vol.seq))

  for(i in seq_along(vol.seq)) {
    setData(base.env, underlying, 'ATMVol', dtCalc, vol.seq[i])
    p[,i] <- getValue(inst, 'Price', dtCalc, base.env)
    d[,i] <- getValue(inst, 'Delta', dtCalc, base.env)
    g[,i] <- getValue(inst, 'Gamma', dtCalc, base.env)
    v[,i] <- getValue(inst, 'Vega', dtCalc, base.env)
  }

  p
  
  old.par <- par(no.readonly = TRUE)
  par(mfrow=c(2,2),oma=c(5,0,0,0),mar=c(2,2,2,1))

  # Price
  plot(und.seq, p[,1], type='n',
       main="Price", xlab="", ylab="")
  topocol <- topo.colors(length(vol.seq))
  for (i in 2:length(vol.seq))
    lines(und.seq, p[,i], col=topocol[i])

  # Delta
  plot(und.seq, d[,1],type='n',
       main="Delta", xlab="", ylab="")

  for (i in 2:length(vol.seq))
    lines(und.seq, d[,i], col=topocol[i])

  # Gamma
  plot(und.seq, g[,1],type='n',
       main="Gamma", xlab="", ylab="")
  for (i in 2:length(vol.seq))
    lines(und.seq, g[,i], col=topocol[i])

  # Vega
  plot(und.seq, v[,1],type='n',
       main="Vega", xlab="", ylab="")
  for (i in 2:length(vol.seq))
    lines(und.seq, v[,i], col=topocol[i])

  mtext(text=paste(inst@desc, "\nrate: 0.03",
          "Underlying from", und.seq[1],"to", und.seq[length(und.seq)],
          "Volatility  from",vol.seq[1], "to",vol.seq[length(vol.seq)]),
        side=1,font=1,outer=TRUE,line=3)
  par(old.par)
  }
```

### Vanilla Option

European calls and puts are called “vanilla” in the banking industry,
presumably due to their simple structures. The payoff of a vanilla
option (call) is:

$$(S_T-K)^+$$ and its value (call) is given by:
$$e^{-rT} \left(F_T N(d_1)-K N(d_2)\right)$$

where $F_T$ is the forward value of the underlying asset at $T$.

For at-the-money options, a simple approximation to the Black-Scholes
price is: $$V = e^{-rT} F_T \frac{\sigma\sqrt{T}}{\sqrt{2\pi}}$$

The formula above makes use of the approximation:

$$N(x) \approx \frac{1}{2} + \frac{x}{\sqrt{2\pi}} + O(x^2)$$

The evolution of price and greeks as a function of the underlying asset
and volatility is shown in figure \[fig:eu-call\]. The graph is obtained
with the following script:

```{r, label=Vanilla, echo=TRUE}
  dtExpiry <- dmy('01jan2011')
  underlying <- 'IBM'; Strike <- 100; K<-1

  b <- fInstrumentFactory("vanilla", quantity=1,
                    params=list(cp='c', strike=Strike,
                    dtExpiry=dtExpiry,
                    underlying=underlying,
                    discountRef='USD.LIBOR', trace=FALSE))

  # define two vectors for the underlying and the volatility
  # all other market data is fixed
  und.seq <- seq(40, 160, by=5)
  vol.seq <- seq(0.1, 0.9, by=0.1)

  base.env <- DataProvider()
  dtCalc <- dmy('01jan2010')
  setData(base.env, underlying, 'DivYield', dtCalc, .02)
  setData(base.env, 'USD.LIBOR', 'Yield', dtCalc, .02)

  OptionDashBoard(b, base.env, dtCalc, und.seq, vol.seq)
```

```{r, label=fig-Vanilla-2,echo=FALSE, fig.width=5, fig.height=5}
  <<Vanilla>>
```

Several observations are worth mentioning:

-   In high volatility conditions, the delta is almost linear, but takes
    on more pronounced a “S” shape as volatility decrease.

-   The gamma is always positive, reflecting the uniformly convex shape
    of the premium function.

-   The vega is also always positive: an increase in volatility always
    results in an increase in premium.

We will see shortly that these last two features have important
implications for risk management.

### Binary Option

A binary option pays a fixed amount if the underlying asset is in the
money at expiry.

The payoff is: $$1_{S_T>K}$$

Recall that $N(d_2)$ is the risk-neutral probability that $S_T>K$ at
expiry. The value of a binary call is therefore:

$$e^{-rT} N(d_2)$$

Alternatively, observe that a binary option struck at $K$ can be
replicated by a very tight call spread:
$$C_{\mbox{B}}(K) = \lim_{h \rightarrow 0} \frac{C_{\mbox{V}} (K-h) - C_{\mbox{V}} (K-h)}{h}$$

where $C_{\mbox{B}}(K)$ is the price of a binary call struk at $K$and
$C_{\mbox{V}}(K)$ the price of a vanilla call. The value of a binary
call is the negative of the derivative of a vanilla call price with
respect to strike.

$$\begin{aligned}
C_{\mbox{B}}(K) &=& -\frac{\partial C_{\mbox{V}} (K)}{\partial K} \\
&=& e^{-rT} N(d_2)\end{aligned}$$

Recall that the derivative of a vanilla call price with respect to spot
is $N(d_1)$, therefore, the shape of the binary price function is
similar to the shape of the delta of a European option, and the shape of
the delta of a binary call is similar to the shape of the gamma of a
vanilla option.

As a consequence, the gamma of a binary option changes sign, and so does
the vega. Options with gamma that changes sign are costly to hedge, as
simulations will demonstrate in section.

The dashboard in Figure \[fig:eu-bin-call\] is generated with the
following script:

```{r, label=Binary, echo=TRUE}
  dtExpiry <- dmy('01jan2011')
  underlying <- 'IBM'; Strike <- 100; K<-1

  bi <- fInstrumentFactory("binary", quantity=1,
                   params=list(cp='c', strike=Strike,
                   dtExpiry=dtExpiry,
                   underlying=underlying,
                   discountRef='USD.LIBOR', trace=FALSE))

  # define two vectors for the underlying and the volatility
  # all other market data is fixed
  und.seq <- seq(40,160,by=5)
  vol.seq <- seq(0.1,0.9,by=0.1)

  base.env <- DataProvider()
  dtCalc <- dmy('01jan2010')
  setData(base.env, underlying, 'DivYield', dtCalc, .02)
  setData(base.env, 'USD.LIBOR', 'Yield', dtCalc, .02)

  OptionDashBoard(bi, base.env, dtCalc, und.seq, vol.seq)
```

```{r, echo=FALSE, fig.width=6, fig.height=6}
  <<Binary>>
```

### Barrier Option

There is a large variety of barrier options. Haug \[[@Haug2006]\]
provides a long list of pricing formulae for standard and complex
barrier payoffs. As an illustration we consider here the “up and out”
call, which is a regular European call that is canceled if the
underlying asset breaches an upper barrier before expiry.

Payoff:

$$(S_T-K)^+1_{\max_t S_t < B}$$

This instruments introduces a new twist in the option landscape, since
this call option can have a negative delta. As for binary options, the
gamma can switch sign, and finally, observe the magnitude of the vega:
the maximum value, in absolute term, is twice as large as for european
options of comparable strike and maturity.

The dashboard in Figure \[fig:eu-ba-call\] is generated with the
following script:

```{r, label=Barrier, echo=TRUE}
  dtExpiry <- dmy('01jan2011')
  underlying <- 'IBM'; Strike <- 100; barrier <- 160

  ba <- fInstrumentFactory("standardbarrier", quantity=1,
                   params=list(cp='cuo', strike=Strike,
                   barrier=barrier, rebate=0, dtExpiry=dtExpiry,
                   underlying=underlying,
                   discountRef='USD.LIBOR', trace=FALSE))

  # define two vectors for the underlying and the volatility
  # all other market data is fixed
  und.seq <- seq(40,160,by=5)
  vol.seq <- seq(0.1,0.9,by=0.1)

  base.env <- DataProvider()
  dtCalc <- dmy('01jan2010')
  setData(base.env, underlying, 'DivYield', dtCalc, .02)
  setData(base.env, 'USD.LIBOR', 'Yield', dtCalc, .02)

  OptionDashBoard(ba, base.env, dtCalc, und.seq, vol.seq)
```

```{r, echo=FALSE, fig.width=6, fig.height=6}
  <<Barrier>>
```

### Asian Option

An asian option is an option on the average value of the underlying
asset, computed over a period of time prior to expiry. This type is very
popular in commodities markets, because it captures the price risk on a
market participant that produces or consumes the asset at a steady rate
over time.

In spite of its apparent complexity, asian options are in fact fairly
easy to manage. One can think of an Asian option as a regular European
option, with an underlying asset that is the average value. The value of
an asian call is lower than the value of a comparable european vanilla
call, since the volatility of an average is lower than the volatility of
the price at expiry. For the same reason, the vega of an Asian option is
about half of the vega of its European conterpart.

The dashboard in Figure \[fig:eu-as-call\] is generated with the
following script:

```{r, label=Asian, echo=TRUE}
  dtExpiry <- dmy('01jan2011')
  underlying <- 'IBM'; Strike <- 100

  ba <- fInstrumentFactory("asian", quantity=1,
                   params=list(cp='c', strike=Strike,
                   dtExpiry=dtExpiry, dtStart=dtCalc, dtEnd=dtExpiry, 
                   underlying=underlying,
                   discountRef='USD.LIBOR', trace=FALSE))

  # define two vectors for the underlying and the volatility
  # all other market data is fixed
  und.seq <- seq(40,160,by=5)
  vol.seq <- seq(0.1,0.9,by=0.1)

  base.env <- DataProvider()
  dtCalc <- dmy('01jan2010')
  setData(base.env, underlying, 'DivYield', dtCalc, .02)
  setData(base.env, 'USD.LIBOR', 'Yield', dtCalc, .02)

  OptionDashBoard(ba, base.env, dtCalc, und.seq, vol.seq)
```

```{r, od-Asian-N,echo=FALSE, fig.width=6, fig.height=6}
  <<Asian>>
```
