# Dynamic Hedging in the Black-Scholes Framework

```{r setup, echo=FALSE, include=FALSE, cache=FALSE}
knitr::opts_chunk$set(
  cache = TRUE,
  dev.args = list(bg = "transparent"),
  tidy=TRUE,
  tidy.opts=list(width.cutoff=45)
)
```

```{r, label=load-DynamicHedging, results='hide'}
  library(fInstrument)
  library(DynamicSimulation)
  library(empfin)
  library(gplots)
  library(BB)
  library(splines)
  library(SuppDists)
  library(moments)
```

In this chapter, we investigate the effectiveness of hedging strategies
in the Black-Scholes framework, and assess the impact of departures from
the assumptions of the Black-Scholes model. Indeed, the ultimate measure
of risk is the quality of the dynamic replicating strategy, and the
distribution of the residual wealth at expiry of the option. Throughout
this chapter, we place ourselves in the situation of a financial
institution that sells a derivative and dynamically hedges the risk.

Self-Financing Replicating Portfolio
------------------------------------

### Continuous rebalancing

Let $C(S_t,t)$ be the price of a call option at time
$t, 0 \leq t \leq T$, with the spot price being $S_t$. We first show
that we can construct a trading strategy involving the underlying asset
and a bank account, which will replicate exactly the option payoff
$\max(S_T - K, 0)$ at time $T$ . This strategy is self-financing,
meaning that it does not require any cash inflow, nor does it generate
any cash outflow after $t = 0$. Because of this self-financing property,
the value of the option must be equal to the initial value of this
replicating portfolio. Otherwise, a sure profit could be gained by
trading the option against the replicating portfolio. The replicating
portfolio $P$ is made of $\alpha_t$ shares and a notional amount
$\beta_t$ of risk-free bond. The portfolio value is thus:

$$P_t = \alpha_t S_t + \beta_t B_t
\label{eq:portfolio-1}$$

where $B_t$ is the price of the risk-free bond at time $t$:

$$B_t = e^{-r(T-t)}$$

and $\alpha_t = \frac{\partial C}{\partial S}$. At $t=0$ the portfolio
has by definition the same value as the option:

$$P_0 = C(S_0 , 0)$$

The residual wealth, $V_t$ is the difference between the replicating
portfolio and the option:

$$V_t = P_t - C(S_t, t)$$

We show that in fact $V_t = 0, \ 0 \leq t \leq T$. Using the definition
of the hedge portfolio, we have:

$$\begin{aligned}
V_t &=& \frac{\partial C}{\partial S} S_t + \beta_t B_t - C(S_t, t) \\
&=& \Pi_t + \beta_t B_t
\beta_t B_t\end{aligned}$$

Differentiate:

$$dV_t = d\Pi_t + \beta_t dB_t$$

From the derivation of the Black-Scholes equation (see Appendix or
@Hull1997), we know that the portfolio $\Pi_t$ is riskless:
$$d\Pi_t = r\Pi dt$$

therefore,

$$\begin{aligned}
d V_t &=& r \Pi_t dt + \beta_t r B dt \nonumber \\
&=& r V_t dt \label{eq:ct-1}\end{aligned}$$

Since $V_0 = 0$, equation \[eq:ct-1\] has a unique solution which is:

$$V_t = 0, \ 0 \leq t \leq T$$

and we conclude, that portfolio $P$ replicates the option at all time,
and in particular at expiry. The dynamic of the replication is governed
by the assumptions of the Black-Scholes framework,which are recalled
here:

-   The underlying asset follows a geometric Brownian motion process,
    with constant and known volatility. The actual drift does not
    matter.

-   The interest rate is constant.

-   Trading is frictionless: there are no trading costs, shares can be
    purchase in arbitrary fractions, trading can take place
    continuously.

Clearly, none of these assumptions are verified in real life, and we
expect that departure from these assumptions will affect the quality of
the dynamic replication described by the Black-Scholes model. The
purpose of this chapter is to investigate the effect of departures from
these assumptions, and address the question: how useful is the
Black-Scholes model in a environment where Black-Scholes assumptions are
not verified?

We will adress this question by means of simulations that are now
described.

### Discreet rebalancing

Since trading does not occur continuously, a replicating strategy must
consist in a decision rule that is implemented at regular intervals. In
the case of the delta-hedging strategy the rule will be:

1.  Define a rebalancing interval $\Delta t$. Trading dates with
    therefore be $\Delta t, 2\Delta t, \ldots, T-\Delta t$.

2.  At each trading date $t$, buy or sell the underlying stock so that
    the hedge portfolio holds a quantity of risky asset equal to
    $\frac{\partial C_t}{\partial S_t}$.

To describe the process step by step, we use the following notation:

$C_t$

:   Value of derivative at time $t$

$V_t$

:   Value of hedge portfolio

$B_t$

:   Value of a pure discount bond maturing at T

$\alpha_t$

:   Delta of derivative, also the quantity of risky asset in the
    replicating portfolio

$\beta_t$

:   Quantity of riskless asset in hedge.

At $t = 0$, the derivative is sold at price $C_0$ and the proceeds are
used to purchase a hedge portfolio. The initial hedge is
$\alpha_0 S_0 + \beta_0 B_0$, where $\beta_0$ is computed from the
acounting identity: $$C_0 = \alpha_0 S_0 + \beta_0 B_0$$

At trading time $t$, the decision rule is as follows:

1.  Compute the value of the hedge portfolio formed at the previous time
    step:

    $$V_t = \alpha_{t-\Delta t} S_t + \beta_{t-\Delta t} B_t$$

2.  Compute the amount of risky security to hold:

    $$\alpha_t = \frac{\partial C_t}{\partial S_t}$$

3.  Since the strategy must be self-financing, determine the quantity of
    bond to be lent or borrowed:
    $$\beta_t = \frac{V_t - \alpha_t S_t}{B_t}$$

At expiry of the derivative, the bond is worth par, and the residual
wealth is:

$$-C_T + \alpha_{T-\Delta t} S_T + \beta_{T-\Delta t}$$

The quality of a model is ultimately measured by the residual error: the
liquidation value of the replicating portfolio, less the exercise value
of the option: $V_T - C_T$.

### Illustration

Let’s illustrate this rebalancing process through a simple example that
will also demonstrate the self-financing nature of the portfolio. A
financial institution writes (sells) an at-the-money option on a stock
that does not pay any dividend, and worth €100. The option expires in
two months, and for illustrative purpose, the hedge will be rebalanced
every week (in practice, of course, dynamic rebalancing is done much
more often). Interest rate is $2\%$ and volatility $30\%$.

```{r, }
   r <- 0.02
   T <- 2/12
   S0 <- 100
   K <- 100
   sigma <- 0.3
   
   p <- GBSOption(TypeFlag = "c", S = S0, X = K, Time = T, r = r,
   b = r, sigma)@price
   delta <- GBSGreeks("delta", TypeFlag = "c", S = S0, X = K, Time = T,
   r = r, b = r, sigma)
```

The initial hedge portfolio is constructed as follows:

-   The value of the hedge portfolio is the value of the derivative:
    $P_0 = `r round(p, 2) `$.

-   The delta of the option is:
    $$\frac{\partial C_0}{\partial S_0} = `r round(delta,2) `$$

    Thus the hedge portfolio must also hold
    $\alpha_0 = `r round(delta,2) `$ units of stocks, for a cost of
    $\alpha_0 \times 100 = `r round(delta*100, 2) `$.

-   this position is funded by a riskless borrowing of
    $\beta_0 B = P_0 - 100 \alpha_0 = 
      -48.48$.

-   Since the zero-coupon bond is worth $B_0 = e^{-rT} = 0.997$, we get
    $\beta_0 = -48.64$.

At every time step, the portfolio is adjusted acording to the algorithm
of the previous section.

This calculation is easily performed with the supplied packages:

```{r, }
   dtStart <- myDate("01jun2010")
   dtObs <- dtStart
   dtExpiry <- myDate("01aug2010")
   underlying <- "IBM"
   K <- 100
   sigma <- 0.3

   a <- fInstrumentFactory("vanilla", quantity = 1, params = list(cp = "c",
   strike = K, dtExpiry = dtExpiry, underlying = underlying,
   discountRef = "USD.LIBOR", trace = FALSE))

   base.env <- DataProvider()

   setData(base.env, underlying, "Price", dtObs, 100)
   setData(base.env, underlying, "DivYield", dtObs, 0.0)
   setData(base.env, underlying, "ATMVol", dtObs, sigma)
   setData(base.env, underlying, "discountRef", dtObs, "USD.LIBOR")
   setData(base.env, "USD.LIBOR", "Yield", dtObs, 0.02)
```

At this stage, we can price the asset with

```{r, }
  p <- getValue(a, "Price", dtStart, base.env)
```

which gives a value of $p = `r round(p,2) `$. Next, define the
parameters to simulate a dynamic hedging policy with weekly rebalancing
and two price paths:

```{r, sim-setup}
  nbSteps <- 8
  nbPaths <- 2
  dtSim <- seq(dtObs, dtExpiry, length.out = nbSteps + 1)
```

The simulated price path is generated by

```{r, sim-setup-2, warning=FALSE}
  tSpot <- pathSimulator(dtSim = dtSim, nbPaths = nbPaths, innovations.gen = sobolInnovations,
  path.gen = logNormal, path.param = list(mu = 0, sigma = sigma),
  S0 = 100, antithetic = F, standardization = TRUE, trace = FALSE)
```

and the simulated paths are placed in a new data provider:

```{r, sim-setup-3}
  sce.env <- DataProvider(parent = base.env)
  setData(sce.env, underlying, "Price", time(tSpot), as.matrix(tSpot))
```

We can now run the delta-hedge strategy along each path:

```{r, delta-hedge-1}
  assets = list(a)
  res <- deltaHedge(assets, sce.env, params = list(dtSim = time(tSpot),
    transaction.cost = 0), trace = FALSE)
```

The result is a data structure that containts the residual wealth
(hedging error) per scenario.

Tables \[tab:dyn-sim-1\] and \[tab:dyn-sim-2\] illustrate how the hedge
portfolios evolve over time, based on the path of the underlying asset.
In Table \[tab:dyn-sim-2\], the option expires out of the money, and the
hedge portfolio ends up being exclusively cash. In
Table \[tab:dyn-sim-1\], the option expires in the money, and the hedge
portfolio holds at expiry one unit of stock. Note the amount of leverage
required by delta hedging when the option expires in the money: the
option payoff is 45 cents, and this is replicated by a portfolio holding
€100.45 worth of stock, funded by borrowing €99.00.

The data shown in both tables are extracted from the simulation results
and formated with the following function.

```{r, sim-results-1, echo=FALSE}
  myTable <- function(iScenario, res, ...) {
  tSpot <- res$spot
  nbSteps <- dim(tSpot)[1]
  z <- matrix(nrow = nbSteps, ncol = 6)
  z[, 1] <- seq(1, nbSteps)
  z[, 2] <- tSpot[, iScenario]
  z[, 3] <- res$stock[, iScenario]
  z[, 4] <- res$price[, iScenario]
  z[, 5] <- res$bond[, iScenario]
  z[, 6] <- res$portfolio[, iScenario]
  colnames(z) <- c("time", "stock price", "delta", "option",
  "bond pos", "hedge port.")
  xtable(z, ...)
  }
  tab.1 <- myTable(1, res)
  tab.2 <- myTable(2, res)
```

```{r, results='asis',echo=F,fig=F}
  print(tab.1, include.colnames=T,floating=FALSE)
```

```{r, results='asis',echo=F,fig=F}
  print(tab.2, include.colnames=T,floating=FALSE)
```

### A More Realistic Example

More time steps and scenarios are needed in order to accurately assess
the distribution of residual wealth. The next experiment uses the
parameters of the previous section, with 200 time steps and 1000
sceanrios.

Figure \[fig:paths-2\] displays a few simulated paths for the underlying
asset, and an histogram of the residual wealth.

```{r, sim-setup-10, cache=TRUE, warning=FALSE}
   nbSteps <- 200
   nbPaths <- 1000
   dtSim <- seq(dtObs, dtExpiry, length.out = nbSteps + 1)
   tSpot <- pathSimulator(dtSim = dtSim, nbPaths = nbPaths, innovations.gen = sobolInnovations,
   path.gen = logNormal, path.param = list(mu = 0, sigma = sigma),
   S0 = 100, antithetic = F, standardization = TRUE, trace = FALSE)
   sim.env <- DataProvider(parent = base.env)
   setData(sim.env, underlying, "Price", time(tSpot), as.matrix(tSpot))
   sim.deltaHedge <- deltaHedge(assets, sim.env, params = list(dtSim = time(tSpot),
   transaction.cost = 0), trace = FALSE)
```

```{r, label=fig-price-plot-2,echo=FALSE,fig.width=6,fig.height=6}
  par(mfrow=c(1,2))
  plot(tSpot[,1:50], plot.type='single', ylab='Price', format="%b %d")
  hist(tail(sim.deltaHedge$wealth,1), 50, xlab="Hedging error", main='')
  par(mfrow=c(1,1))
```

As expected, the distribution of residual wealth is centered, and has a
standard deviation of

```{r, sim-calc-11}
  sd.wealth <- sd(t(tail(sim.deltaHedge$wealth, 1)))
```

A confidence interval (say, 80%) for the P&L of the delta-hedging
strategy is readily computed:

```{r, sim-calc-2}
  pl.confidence <- qnorm(0.2, mean = 0, sd = sd.wealth)
```

which gives a value of $`r round(pl.confidence,2) `$. In other words,
there is a 80% probability that this delta-hedging strategy will yield a
P&L greater than $`r round(pl.confidence,2) `$ per € of nominal. In
the next couple of sections, we will provide a more rigorous analysis of
these observations.

With this simulation tool, we will now consider the effect on hedging
effectiveness of:

-   discrete rather than continuous hedging,

-   transaction costs, and

-   unknown and stochastic volatility.

Hedging Error due to Discrete Hedging
-------------------------------------

We first consider the risk related to discreet hedging. The dynamic of
the replicating portfolio, $V_t$ rebalanced at times
$t_i, i= 1, \ldots, N$. For $t \in [t_i, t_{i+1})$ is:

$$dV_t = \frac{\partial C(t_i, S_{t_i})}{\partial S} dS_t
\label{eq:dV}$$

The dynamic of the option price is:

$$dC_t = \frac{\partial C(t, S_{t})}{\partial S} dS_t$$

The difference between the two processes being that in the hedge
portfolio, the quantity of stock
$\sfrac{\partial C(t_i, S_{t_i})}{\partial S_{t_i}}$ (Eq. \[eq:dV\]), is
held constant between two rebalancing dates.

The hedging error process $\epsilon_t = C_t - V_t$ is therefore:

$$d\epsilon_t = \left(\frac{\partial C(t, S_{t})}{\partial S} - \frac{\partial C(t_i, S_{t_i})}{\partial S} \right) dS_t$$

It can be shown that:

$$\varinjlim_{h \rightarrow 0} E^Q \left[ \frac{\epsilon_T^2}{\Delta t} \right] = E^Q \left[ \frac{1}{2} \int_0^T \left( \frac{\partial^2 C(S_t,t)}{\partial S_t^2} \right)^2 S^4_t \sigma^4 dt\right]$$

It can also be shown that the process $\epsilon_t$ converges in
distribution to a process that has the following expression:

$$\frac{\epsilon_t}{\sqrt{\Delta t}} \rightarrow \frac{1}{\sqrt{2}} \int_0^t \frac{\partial^2 C(S_u, u)}{\partial S^2} \sigma^2 S^2_u dZ_u$$

This expression highlights the following points:

1.  The variance of the hedging is inversely related to the hedging
    frequency, and

2.  is directly related to the magnitude of
    $\frac{\partial^2 C(S_u, u)}{\partial S^2}$, which is the Gamma of
    the option.

### Empirical Test

To test the impact of hedging frequency on the distribution of
delta-hedging error, we simulate the dynamic delta hedging of a ATM
option. The price paths are generated with the same volatility as the
one used for pricing, and there are no transaction costs. Therefore, the
only source of hedging error is the discrete hedging policy. The
simulation involves several steps that are outlined next.

First, define the derivative to be dynamically hedged:

```{r, label=hf-1, echo=TRUE}
  dtExpiry <- mytDate('01jan2011')
  underlying <- 'IBM'
  K<-100
  a <- fInstrumentFactory("vanilla", quantity=1,
                    params=list(cp='c', strike=K,
                      dtExpiry=dtExpiry, 
                      underlying=underlying,
                      discountRef='USD.LIBOR', trace=FALSE))
```

Next, we define the market data that we be used in the simulation. The
base environment holds data that will not change from one simulation to
the next.

```{r, label=hf-2, echo=TRUE}
  dtStart <- mytDate('01jan2010')
  nbPaths <- 1000;
  sigma <- .3

  base.env <- DataProvider()
  setData(base.env, underlying, 'Price', dtStart, 100)
  setData(base.env, underlying, 'DivYield', dtStart, .02)
  setData(base.env, underlying, 'ATMVol', dtStart, sigma)
  setData(base.env, underlying, 'discountRef', dtStart, 'USD.LIBOR')
  setData(base.env, 'USD.LIBOR', 'Yield', dtStart, .02)
```

We next perform a series of dynamic hedging simulations, while varying
the number of time steps.

```{r, label=hf-3, echo=TRUE, cache=TRUE, warning=FALSE}
  nb.range <- seq(10, 500, by=50)
  mean.error <- 0.0 * nb.range
  sd.error <- 0.0 * nb.range

  for(i in seq_along(nb.range)) {

  nbSteps <- nb.range[i]

  dtSim <- seq(as.timeDate(dtStart), as.timeDate(dtExpiry),
               length.out=nbSteps+1)

  # price paths 
  tSpot <- pathSimulator(dtSim = dtSim, nbPaths=nbPaths, 
      innovations.gen=sobolInnovations, path.gen=logNormal, 
      path.param = list(mu=0, sigma=sigma), S0=100, antithetic = F, 
      standardization = TRUE, trace = FALSE)

  # derived environment for scenario analysis

  sce.env <- DataProvider(parent=base.env)
  setData(sce.env, underlying, 'Price',
          time(tSpot), as.matrix(tSpot))

  # simulate a delta-hedge strategy along each path

  assets = list(a)
  res <- deltaHedge(assets, sce.env,
                    params=list(dtSim=time(tSpot),
                    transaction.cost=0), trace=FALSE)

  expiry.error <- tail(res$wealth,1)

  mean.error[i] <- mean(expiry.error)
  sd.error[i] <- sd(as.vector(expiry.error)) 
  }
```

Figure \[fig:delta-steps\] illustrates the mean and standard deviation
of the hedging error as a function of the number of rehedging dates. The
figure also shows the curve $$y = a + b \frac{1}{\sqrt{N}}$$

fitted to the standard deviation of the hedging error. The theory
predicts that the standard deviation of the hedging error is
proportional to $\frac{1}{\sqrt{N}}$, where $N$ is the number of
re-hedging dates.

```{r, label=Delta-Hedge, echo=TRUE}
  plotCI(x=nb.range, y=mean.error, uiw=sd.error/2, liw=sd.error/2, ylim=c(min(mean.error-sd.error/2), 
         max(mean.error+sd.error/2)), 
         xlab='Nb of rebalancings', ylab='Hedging error')

  q <- lm((mean.error+sd.error/2) ~ sqrt(1/nb.range))
  yy <- predict(interpSpline(nb.range, q$fitted.values))
  # lines(nb.range, q$fitted.values, lwd=2, col='red')
  lines(yy, lwd=2, col='red')
```

```{r, label=fig-Delta-Hedge, echo=FALSE, fig.width=5, fig.height=5}
  <<Delta-Hedge>>
```


To investigate the effect of gamma on hedging error, we will compare the
effectiveness of delta hedging, applied to two options that initially
have the same price and delta. The initial Gammas are within 10% of each
other. However, the options differ substantially in one aspect: the
first option is a European call, and its Gamma varishes when the option
gets deep in the money, while the second option is a binary call, whose
Gamma remains large and actually changes sign under the same conditions.

To perform the experiment, we first define the data environment: the
spot price is 100, option maturity will be 1 year.

```{r, label=data-env, echo=TRUE}
  dtExpiry <- myDate('01jan2011')
  dtStart <- myDate('01jan2010')
  underlying <- 'IBM'
  sigma <- .3

  base.env <- DataProvider()
  setData(base.env, underlying, 'Price', dtStart, 100)
  setData(base.env, underlying, 'DivYield', dtStart, .02)
  setData(base.env, underlying, 'ATMVol', dtStart, sigma)
  setData(base.env, underlying, 'discountRef', dtStart, 'USD.LIBOR')
  setData(base.env, 'USD.LIBOR', 'Yield', dtStart, .02)
```

The strike of the European option is set to $K=130$.

```{r, label=op.eu, echo=TRUE}
  K <- 130
  op.eu <- fInstrumentFactory("vanilla", quantity=1,
                    params=list(cp='c', strike=K,
                      dtExpiry=dtExpiry, 
                      underlying=underlying,
                      discountRef='USD.LIBOR', trace=FALSE))

  P <- getValue(op.eu, 'Price', dtStart, base.env)
  D <- getValue(op.eu, 'Delta', dtStart, base.env)
```

The price is $P=`r round(P,2) `$, delta is $D=`r round(D,2) `$.

The strike and payoff of the binary option are set in order to match the
price and delta of the European call, so as to make a meaningful
comparison of hedging error. This involves solving a set of two
non-linear equations (price and delta) in two unknown (payoff and
strike). These equations are solved by means of the following function:

```{r, label=vb-1, echo=TRUE, message=FALSE}
  price.delta.error <- function(x, base.env) {
   Q <- x[1]
   K.b <- x[2]
   
   op.bi <- fInstrumentFactory("binary", quantity=Q,
                    params=list(cp='c', strike=K.b,
                      dtExpiry=dtExpiry, 
                      underlying=underlying,
                      discountRef='USD.LIBOR', trace=FALSE))

   Pa <- getValue(op.bi, 'Price', dtStart, base.env)
   Da <- getValue(op.bi, 'Delta', dtStart, base.env)

   f <- rep(NA, 2) 
   f[1] <- (P-Pa) 
   f[2] <- (D-Da)
   f
  }

  ans <- dfsane(par=c(100, 100), fn=price.delta.error, base.env=base.env)

  Q <- ans$par[1]
  K.b <- ans$par[2]
```

The solution is a binary call with payoff $Q=`r round(Q,2) `$ and
strike $K=`r round(K.b,2) `$.

```{r, label=op.bi, echo=TRUE}
  op.bi <- fInstrumentFactory("binary", quantity=Q,
                    params=list(cp='c', strike=K.b,
                      dtExpiry=dtExpiry, 
                      underlying=underlying,
                      discountRef='USD.LIBOR', trace=FALSE))

  G.bi <- getValue(op.bi, 'Gamma', dtStart, base.env)
  G.eu <- getValue(op.eu, 'Gamma', dtStart, base.env)
```

The Gammas of the two options are not identical, but within 10% of each
others: $`r round(G.eu, 4) `$ for European option,
$`r round(G.bi,4) `$ for the binary option.

To illustrate the similarity between the two derivatives, we plot in
Figure\[fig:eu-vanilla-binary\] the prices of both options for a range
of underlying values, with a zoom around the current spot price.

```{r, label=VanillaBinary, echo=TRUE}
  sce.1 <- t(as.matrix(seq(60, 160, length.out=50)))
  sce.2 <- t(as.matrix(seq(60, 350, length.out=50)))

  # derived data provider with price scenarii
  sim.env.1 <- DataProvider(parent=base.env)
  setData(sim.env.1, underlying, 'Price', dtStart, sce.1)

  sim.env.2 <- DataProvider(parent=base.env)
  setData(sim.env.2, underlying, 'Price', dtStart, sce.2)

  # Compute and plot NPV per scenario for underlying spot price
  p.eu.1 <- getValue(op.eu, 'Price', dtStart, sim.env.1)
  p.bi.1 <- getValue(op.bi, 'Price', dtStart, sim.env.1)

  p.eu.2 <- getValue(op.eu, 'Price', dtStart, sim.env.2)
  p.bi.2 <- getValue(op.bi, 'Price', dtStart, sim.env.2)

```

```{r, label=fig-VanillaBinary,echo=FALSE, fig.width=5, fig.height=5}
  par(mfrow=c(1,2))
  plot(sce.1, p.eu.1, ylim=c(min(c(p.eu.1, p.bi.1)), max(c(p.eu.1, p.bi.1))),
       type='l', lwd=2, ylab='Price', xlab='Spot', col='green')
  lines(sce.1, p.bi.1, type='l', lwd=2, col='red') 

  plot(sce.2, p.eu.2, ylim=c(min(c(p.eu.2, p.bi.2)), max(c(p.eu.2, p.bi.2))),
       type='l', lwd=2, ylab='Price', xlab='Spot', col='green')
  lines(sce.2, p.bi.2, type='l', lwd=2, col='red') 
  par(mfrow=c(1,1))

```

A delta hedging strategy is next simulated for both derivatives, using
100 steps, 5000 paths.

```{r, label=va-bi-sim-1, echo=TRUE}
  nbSteps <- 100;
  nbPaths <- 5000;

  dtSim <- seq(as.timeDate(dtStart), as.timeDate(dtExpiry),
               length.out=nbSteps+1)
```

```{r, label=va-bi-sim-2, echo=TRUE, cache=TRUE, warning=FALSE}
  # price paths 
  tSpot <- pathSimulator(dtSim = dtSim, nbPaths=nbPaths, 
      innovations.gen=sobolInnovations, path.gen=logNormal, 
      path.param = list(mu=0, sigma=sigma), S0=100, antithetic = F, 
      standardization = TRUE, trace = FALSE)

  # derived environment for scenario analysis
  sce.env <- DataProvider(parent=base.env)
  setData(sce.env, underlying, 'Price',
          time(tSpot), as.matrix(tSpot))

  assets = list(op.eu)
  res1 <- deltaHedge(assets, sce.env,
                    params=list(dtSim=time(tSpot),
                    transaction.cost=0), trace=FALSE)

  assets = list(op.bi)
  res2 <- deltaHedge(assets, sce.env,
                    params=list(dtSim=time(tSpot),
                    transaction.cost=0), trace=FALSE)
```

Figure \[fig:vanilla-binary-error\] displays the histograms of residual
errors.

```{r, label=VB-Histo, echo=T, eval=FALSE}
  par(mfrow=c(1,2))
  # distribution of wealth at horizon
  hist(tail(res1$wealth,1), 50, xlab="wealth", main='Vanilla Call')
  hist(tail(res2$wealth,1), 50, xlab="wealth", main='Binary Call')
  par(mfrow=c(1,1))
```

```{r, label=fig-VanillaBinaryError,echo=FALSE, fig.width=5, fig.height=5}
  <<VB-Histo>>
```

Statistics on the hedging error at expiry are summarized in
Table \[tab:euro-bin\]. The standard deviation of the binary hedging
error is about three times larger than for the vanilla option. Most
importantly, the excess kurtosis is almost 10 times larger, indicating
the possibilily that applying a standard delta-hedging strategy to a
binary option can result in extreme losses (or gains!). All of this in a
context where the only departure from the Black-Scholes assumptions is a
discrete rather than continuous hedging. We next investigate the impact
of transaction costs.

```{r, VB-stats, echo=TRUE}
  w1 <- as.vector(tail(res1$wealth,1))
  w2 <- as.vector(tail(res2$wealth,2))
  df.tmp <- data.frame(Vanilla=w1, Binary=w2)
  funs <- list(mean, stdev, kurtosis, skewness)
  res <- sapply(funs, mapply, df.tmp)
  colnames(res) <- c("mean", "$\\sigma$", "skewness", "kurtosis")
  xm <- xtable(res)
```

```{r, results='asis',echo=F,fig=F}
  print(xm, sanitize.text.function=function(x) {x})
```

Hedging Error due to Transaction Cost
-------------------------------------

Let $S_t$ be the mid price of the underlying asset and $k$ the
proportional transaction cost: the hedger buys at the ask price:
$$S_t^a = S_t (1+\frac{k}{2})$$

and sells at the bid: $$S_t^a = S_t (1-\frac{k}{2})$$

Clearly, transaction costs increase the cost of replication. To
compensate for that, one needs to start with a higher budget, that is,
the option seller should change a higher option premium than the one
given by the Black-Scholes model. This higher option premium can be
obtained in the Black-Scholes framework by increasing volatility.
@Leland1985 shows that one can adjust the Black-Scholes volatility as
follows:

$$\sigma^{+2} = \sigma^2 \left( 1+\sqrt{\frac{2}{\pi}} \frac{k}{\sigma \sqrt{\Delta t}}\right)$$

where $\sigma$ is the Black-Scholes volatility, and $\sigma+$ the
volatility adjusted for transaction costs.

### Empirical Test

To test Leland’s formula, we simulate the delta-hedging of a
at-the-money European option, struck at 100, for transaction costs $k$
ranging from $0$ to $1\%$. The simulation is performed as follows:

```{r, label=Leland-1, echo=TRUE, cache=TRUE}
  tr.range <- seq(0, 1, .1)/100
  nbSteps <- 250

  dtSim <- seq(as.timeDate(dtStart), as.timeDate(dtExpiry),
               length.out=nbSteps+1)

  dt <- as.double(dtExpiry-dtStart)/(365*nbSteps)

  # price paths 
  tSpot <- pathSimulator(dtSim = dtSim, nbPaths=nbPaths, 
      innovations.gen=sobolInnovations, path.gen=logNormal, 
      path.param = list(mu=0, sigma=sigma), S0=100, antithetic = FALSE, 
      standardization = TRUE, trace = FALSE)

  # derived environment for scenario analysis
  sce.env <- DataProvider(parent=base.env)
  setData(sce.env, underlying, 'Price',
          time(tSpot), as.matrix(tSpot))

  mean.error <- 0.0 * tr.range
  sd.error <- 0.0 * tr.range

  for(i in seq_along(tr.range)) {

  # simulate a delta-hedge strategy along each path

  assets = list(a)
  res <- deltaHedge(assets, sce.env,
                    params=list(dtSim=time(tSpot),
                    transaction.cost=tr.range[i]/2), trace=FALSE)

  expiry.error <- tail(res$wealth,1)

  mean.error[i] <- mean(expiry.error)
  sd.error[i] <- sd(as.vector(expiry.error)) 
  }
```

We next compare the expected hedging error to the value predicted by
Leland’s formula. To do so, we convert the expected hedging error into a
volatility adjustment, using the first order approximation:

$$P(\sigma) = P(\sigma^*) + (\sigma-\sigma^*) \frac{\partial P}{\partial \sigma}$$

where $P(\sigma)$ is the price of the option with volatility $\sigma$.

```{r, label=Leland-2, echo=TRUE}
  sigma.leland <- sigma * sqrt(1+ sqrt(2/pi) * (tr.range/(sigma*sqrt(dt))))

  vega <- getValue(a, 'Vega', dtStart, base.env)
  sigma.sim <- sigma - mean.error/vega
```

The results are presented in Figure \[fig:delta-leland\]. We find very
good agreement between the empirical result and Leland’s formula.

```{r, label=Leland-plot, echo=FALSE}
  par(mfrow=c(1,2))

  plotCI(x=tr.range, y=mean.error, uiw=sd.error/2, liw=sd.error/2, 
  ylim=c(min(mean.error-sd.error/2), max(mean.error+sd.error/2)), 
  xlab='Round trip cost', ylab='Hedging error')

  plot(x=tr.range, y=sigma.leland, type="b", pch=22, col="blue", lty=2, xlab="Round trip cost", ylab="Volatility")
  lines(x=tr.range, y=sigma.sim, type="b", pch=23, col="green", lty=2, xlab="", ylab="")
  # legend('top', c("Leland", "Simulation"), lty=c(1,1), col=c("blue", "green"))
  par(mfrow=c(1,1))
```

```{r, label=fig-Leland,echo=FALSE, fig.width=5, fig.height=5}
  <<Leland-plot>>
```

Hedging Error due to Unknown Volatility
---------------------------------------

The strongest assumption of the Black-Scholes model is arguably the
assumption that volatility is known and constant. we next investigate
the impact of unknown volatility of hedging error, following a landmark
paper by @Karoui1998.

To simplify the notation, assume now that hedging is done continuously,
but that the volatility of the underlying asset is unknown. The dynamic
of the underlying asset is:

$$dS_t = \mu dt + \sigma_t dW_t$$

The derivative is priced with an estimated volatility $\Sigma$.

The dynamic of the hedge portfolio is:

$$dV_t = \frac{\partial C_t^\Sigma}{\partial S_t} dS_t$$

By Itô’s lemma:

$$dC(S_t, t) = \left( \frac{\partial C}{\partial t} + \frac{1}{2} \frac{\partial^2 C}{\partial S^2} \sigma_t^2 S_t^2 \right) dt + \frac{\partial C}{\partial S} dS_t$$

The price of the derivative verifies the Black-Scholes EDP:

$$\frac{\partial C}{\partial t} + \frac{1}{2} \Sigma^2 S_t^2 \frac{\partial^2 C}{\partial S^2} = 0$$

this yields,

$$d \epsilon_t = dV_t - dC_t = \frac{1}{2} \left[ \Sigma^2 - \sigma_t^2 \right] \frac{\partial^2 C}{\partial S^2} S^2 dt$$

and the hedging error at expiry is thus,

$$\epsilon_T = \frac{1}{2} \int_0^T \left[ \Sigma^2 - \sigma_t^2 \right] \frac{\partial^2 C}{\partial S^2} S^2 dt$$

We identify three components in the above equation:

1.  the nature of the option, characterized by its Gamma.

2.  the behavior of the market, characterized by $\sigma_t$,

3.  the model calibration, summarized here by the Black-Scholes
    volatility $\Sigma$

If the Gamma keeps a constant sign, we can compute the fair value of the
Black-Scholes volatility that sets the expected hedging error to 0 is

$$\sigma_{BS}^2 = \frac{E\int_0^T \Gamma S^2 \frac{\Delta S}{S}^2}{E\int_0^T \Gamma_{BS} S^2 dt}$$

Equation (\[eq:ElKaroui-JB-S\]) also shows that, when the gamma keeps a
constant sign, we can choose a pricing volatility $\Sigma$ such that
$E(\epsilon_T)$ is positive.

Conclusion
----------

The experiments of this chapter highlight the crucial role of Gamma in
dynamic hedging. In short, the original Black-Scholes framework remains
adapted for options for which the Gamma does not change sign. This is
summarized in Table \[tab:bs-adapted\]. In such case, a positive
expected hedging error can be ensured by setting the pricing volatility
sufficiently high or low, as the case may be. Moreover, the Gamma of
these options often vanishes in deep in the money or out of the money
scenarios, and this renders of course the delta hedging strategy very
effective. This may explain why the Black-Scholes model remains so
widely used, despite its well documented shortcomings. Next chapter
describes a framework for dealing with Gamma risk, while remainning in
the Black-Scholes framework.

  Payoff                Is BS Adapted?
  -------------------- ----------------
  Vanilla                    Yes
  European Binary             No
  Up and Out Barrier          No
  Asian                      Yes

Second-Order Hedging
--------------------

In the previous section, we have determined that hedging error is
determined by three factors: the hedging frequency, the magnitude of the
option gamma, and finally the difference between $\Sigma$ (the BS
volatility used for pricing and hedging) and the volatility that is
actually experienced, $\sigma_t$.

In this section, we investigate strategies for controlling the risk
associated with the magnitude of the gamma.
