# The implied volatility

```{r, label=iv-setup, echo=FALSE, results='hide'}
  library(fOptions)
  library(oce)
  library(empfin)
  data(LiborRates)
  data(cl_opt_042109)
```

Listed options are quoted in price. Implied volatility is the volatility
that recovers that quoted price, using the Black-Scholes model.
Computing implied volatility amounts to inverting the BS formula, given
an option price. This chapter discusses the issues involved in such
calculation, and how to represent the result.

Market Data
-----------

Option settlement prices are published every day by exchanges.
Figure \[fig:iv-data\] shows a sample of the data file published every
day by the NYMEX.

### Option data: Settlement prices of options on the Feb09 futures contract

``` {fontsize="\tiny"}
                NEW YORK MERCANTILE EXCHANGE
          NYMEX OPTIONS CONTRACT LISTING FOR 12/29/2008

                           TODAY'S  PREVIOUS  ESTIMATED   
--------CONTRACT--------    SETTLE    SETTLE     VOLUME   
LC    02 09    P   30.00       .53       .85          0    
LC    02 09    P   35.00      1.58      2.28          0    
LC    02 09    P   37.50      2.44      3.45          0    
LC    02 09    C   40.00      3.65      2.61         10    
LC    02 09    P   40.00      3.63      4.90          0    
LC    02 09    P   42.00      4.78      6.23          0    
LC    02 09    C   42.50      2.61      1.80          0    
LC    02 09    C   43.00      2.43      1.66          0    
LC    02 09    P   43.00      5.41      6.95        100    
```

NEW YORK MERCANTILE EXCHANGE NYMEX OPTIONS CONTRACT LISTING FOR
12/29/2008

TODAY’S PREVIOUS ESTIMATED ——–CONTRACT——– SETTLE SETTLE VOLUME LC 02 09
P 30.00 .53 .85 0 LC 02 09 P 35.00 1.58 2.28 0 LC 02 09 P 37.50 2.44
3.45 0 LC 02 09 C 40.00 3.65 2.61 10 LC 02 09 P 40.00 3.63 4.90 0 LC 02
09 P 42.00 4.78 6.23 0 LC 02 09 C 42.50 2.61 1.80 0 LC 02 09 C 43.00
2.43 1.66 0 LC 02 09 P 43.00 5.41 6.95 100

Notice that settlement prices are provided even for options that have
not been traded today. Providing a price for all strikes and maturities
is necessary in order to compute margin calls on open positions. The
exchange uses an interpolation method for estimating these settlement
prices in the absence of actual transactions.

Calculation of implied volatility
---------------------------------

Given an observed price $C^*$, compute the volatility $\sigma$ such
that:

$$\begin{aligned}
 C^* & = & C(S, K, T, r, \sigma) \\
 & = & C(\sigma)\end{aligned}$$

Observe that there is a change in convexity in $f(\sigma)$:

$$\frac{\partial C}{\partial \sigma} = S n(d_1) \sqrt{T}$$

$$\frac{\partial^2 C}{\partial \sigma^2} = S\sqrt{T} n(d_1) \frac{1}{\sigma} \left[  \frac{1}{\sigma^2 T} \ln(\frac{F}{K})^2-\frac{1}{4} \sigma^2 T \right]$$

with $F = Se^{rT}$.

Thus, $C(\sigma)$ is convex on the interval
$(0, \sqrt{\frac{2|\ln(F/K)|}{T}}]$, and concave otherwise.

### Newton’s method

Let $C^*$ be the price of a call option. We look for the volatility
$\sigma^*$ such that $$C^* = C(S, K, T, r, \sigma^*)$$

To ensure convergence of Newton’s method, one must carefully choose the
initial point.

Let $f$ be defined on the interval $[a,b]$ and assume that:

1.  $f(x^*) = 0$ for some $x^* \in [a, b]$

2.  $f'(x) > 0$

3.  $f''(x) \geq 0$

Then Newton’s method converges monotonically from $x_0 = b$. If

1.  $f(x^*) = 0$ for some $x^* \in [a, b]$

2.  $f'(x) > 0$

3.  $f''(x) \leq 0$

Then Newton’s method converges monotonically from $x_0 = a$.

We prove the first assertion. For all $x_n \in [a, b]$,
$$f(x^*) = f(x_n) + (x^* - x_n) \frac{\partial f}{\partial x} + K$$ with
$K > 0$. Therefore, $$x^* < x_n - \frac{\partial f}{\partial x}$$

or,

$$x^* <= x_{n+1} < x_n$$

Newton’s method generates a monotonic sequence that must converge to
$x^*$.

Consider now Newton’s method started at
$$\sigma_0 = \sqrt{\frac{2|\ln(F/K)|}{T}}$$

If $f(\sigma_0)>0$, we are in case I of the above theorem, and Newton’s
method generates a decreasing sequence that converges to $\sigma^*$. If
$f(\sigma_0)<0$ we are in case II, and Newton’s method also converges.
The algorithm is as follows:

### Implied Volatility by Newton’s Method

The following algorithm generates a monotonic series $(\sigma_n)$:

1.  Set $\sigma_0 = \sqrt{\frac{2|\ln(F/K)|}{T}}$

2.  While $|C(\sigma_n) - C^*| > \epsilon$:

    1.  Let
        $$\sigma_{n+1} = \sigma_n + \frac{C^*-C(\sigma_n)}{\frac{\partial C}{\partial \sigma}}$$

    2.  $n \leftarrow n+1$

The algorithm is implemented as follows:

```{r, label=ImpliedVolNewton, echo=T}
  ImpliedVolNewton <- function(p, TypeFlag, S, X, Time, r, b, tol, maxiter=50) {
    F <- S * exp((r-b)*T)
    s <- sqrt(2*log(F/X)/T)
    not_converged <- T
    vega <- GBSGreeks(Selection="vega", TypeFlag, S, X, Time, r, b, s)
    i <- 1
    while(not_converged & (i<maxiter)) {
      err <- (p-GBSOption(TypeFlag, S, X, Time, r, b, s)@price)
      s <- s + err/vega
      not_converged <- (abs(err/vega) > tol)
      i <- i+1
    }
  s
  }
```

A timing test demonstrate the advantage of choosing the initial point
described above:

```{r, label=iv-timing-test-1, echo=T}
  TypeFlag <- 'c'
  S <- 100
  X <- 100
  Time <- 1
  r <- .03
  b <- .01
  sigma <- .314
  tol <- 1e-6

  p <- GBSOption(TypeFlag, S, X, Time, r, b, sigma)@price
```

We perform 100 replications of the same calculation.

```{r, label=iv-timing-test-2, echo=T}
  t1 <- function(n=100) {
  for (i in 1:n) {
  si <- GBSVolatility(p, TypeFlag, S, X, Time, r, b, tol=tol, maxiter=50)
  }}

  t2 <- function(n=100) {
  for (i in 1:n) {
  si <- ImpliedVolNewton(p, TypeFlag, S, X, Time, r, b, tol) }}
```

Statistics for the default implied volatility calculation are:

```{r, label=iv-timing-test-3, echo=T}
  system.time(t1(100))
```

and with Newton’s method and the initial point computed as above:

```{r, label=iv-timing-test-4, echo=T}
  system.time(t2(100))
```

### Secant Method

For deep in the money or out of the money options, where vega is very
small, a solution method that does not require a derivative may be
prefered. The secant method starts with a bracking interval around the
solution, and progressively shrinks it. The algorithm is as follows:

It implementation is straight-forward:

```{r, label=iv-secant-1, echo=T}
  ImpliedVolSecant <- function(p, TypeFlag, S, X, Time, r, b, tol, sBounds, maxiter=100) {
  sMin <- min(sBounds)
  sMax <- max(sBounds)
  pMin <- GBSOption(TypeFlag, S, X, Time, r, b, sMin)@price
  pMax <- GBSOption(TypeFlag, S, X, Time, r, b, sMax)@price
   
  not_converged <- abs(pMin-pMax) > tol
  i <- 1

  while(not_converged & (i<maxiter)) {
    sStar <- (sMin + sMax)/2
    pStar <- GBSOption(TypeFlag, S, X, Time, r, b, sStar)@price
    if(pStar < p) {
      pMin <- pStar;
      sMin <- sStar
    } else {
      pMax <- pStar;
      sMax <- sStar
    }

    not_converged <- (abs(pMin-pMax) > tol)
    i <- i+1
  }

  (sMin+sMax)/2
  }   

  t3 <- function(n=100) {
  for (i in 1:n) {
  si <- ImpliedVolSecant(p, TypeFlag, S, X, Time, r, b, tol, c(.1, 1))
  }
  si
  }
```

A timing test shows the value of this algorithm for a deep our of the
money call:

```{r, label=iv-secant-2, echo=T}
  S <- 100
  X <- 20
  sigma <- .514
  p <- GBSOption(TypeFlag, S, X, Time, r, b, sigma)@price
  n <- 100

  system.time(t1(n))
  system.time(t2(n))
  system.time(t3(n))
```

### Jaeckel’s method

Peter Jackel ([@Jackel2006]) proposes a variant of the above method to
regularize Newton’s method.

Given an option price $p$, we must solve for $\sigma$

$$p = \delta \theta \left[ F \Phi \left(\theta \left[ \frac{\ln(F/K)}{\sigma} + \frac{\sigma}{2} \right] \right) - K \Phi \left(\theta \left[ \frac{\ln(F/K)}{\sigma} - \frac{\sigma}{2} \right] \right) \right]$$

where:

$\delta$

:   discount factor

$\theta$

:   1 for call, -1 for put

$F$

:   Forward price: $S e^{(r-d)T}$

$\sigma$

:   $\sigma \sqrt{T}$

Set:

$$\begin{aligned}
    x & = &\ln(F/K) \\
    b & = & \frac{p}{\delta \sqrt{FK}}\end{aligned}$$

The Black-Scholes equation becomes:

$$b = \theta \left[ e^{x/2} \Phi \left(\theta \left[ \frac{x}{\sigma} + \frac{\sigma}{2} \right] \right) - e^{-x/2} \Phi \left(\theta \left[ \frac{x}{\sigma} - \frac{\sigma}{2} \right] \right) \right]$$

### Normalized Call price as a function of $\sigma$

![image](Graphics/iv-1.pdf){width=".7\textwidth"}

![image](Graphics/iv-1.pdf)

As in the previous section, the normalized price function changes
convexity at $\sigma_c = \sqrt{|x|}$

Jackel further defines $$f(\sigma) = \left\{
        \begin{array}{rr}
            \ln(\frac{b-i}{\overline{b}-i}) & \mbox{if} \overline{b} < b_c \\
            b-\overline{b} & \mbox{otherwise}
        \end{array}
    \right.$$

with:

$b_c$

:   $b(x, \sigma_c, \theta)$

$i$

:   normalized intrinsic value:
    $1_{\theta x>0} \theta \left( e^{x/2} - e^{-x/2}\right)$

$\overline{b}$

:   target normalized price

The function $f(\sigma)$ is now monotonously concave, and the implied
volatility problem amounts to solving: $$f(\sigma) = 0$$

### Transformed objective function

![image](Graphics/iv-2.pdf){width=".7\textwidth"}

![image](Graphics/iv-2.pdf)

### Impact of Dividends

For indices: watch for dividends impact, compute implied dividend rate
from C/P parity, then compute implied vol.
