# Elementary Fixed Income Calculations {#fi:elementary}

```{r 401-BasicIR,echo=FALSE,results='hide'}
  library(timeDate)
  library(empfin)
  library(tufte)
```

```{block, type="epigraph"}
An honest man’s word is as good as his bond
--- Miguel de Cervantes 
```

`r newthought("This")` chapter provides a elementary survey of fixed income calculations,
with the aim of developing a basic intuition regarding the relationship
between the price and the yield of fixed income instruments.

## Bond Pricing

In a simplified framework with even intervals between payments, the
present value of a bond with coupon rate $c$ and maturing in $n$ years
is given by $$P(r) = \sum_{i=1}^n \frac{cN}{(1+r)^i} + \frac{N}{(1+r)^n}
\label{eq:fixedinBonds-1}$$

where:

$c$

:   coupon rate, in decimal form

$N$

:   nominal

$n$

:   maturity, measured in years

$r$

:   yield, in decimal form

The formula can be implemented in one line:

```{r 401-100}
  SimpleBondPrice <- function(coupon, n, N=100, yield) {
    N*(sum(coupon/(1 + yield)^(1:n)) + 1/(1+yield)^n)
  }
```

As an illustration, consider a bond with 3 years to maturity, a $5\%$
coupon rate. With yield at $4\%$, price per $100\$$ nominal amount is

```{r 401-110}
  P <- SimpleBondPrice(coupon=.05, n=3, yield=.04)
```

or $P = `r round(P,2) `$.

## The Price-Yield Curves

Figure \[fig:PY-Coupon\] represents the price-yield relationship for 20
year bonds with coupons ranging from 0 to 15%. In addition to the
obvious non-linear nature of the price-yield relationship, it is worth
noting that the convexity of the price-yield curve increases with the
coupon rate.

```{r, label=PY-Coupon, echo=T}
  ## plot price-yield as a function of coupon
  cp <- c(0, .05, .1, .15)
  cl <- c('red', 'green', 'blue', 'black')
  y <- seq(0, .20, .01)
  n <- 20
  p <- matrix(0, length(cp), length(y))
  for(i in seq_along(cp)) {
    p[i,] <- sapply(y, function(y) SimpleBondPrice(cp[i], n, yield=y))
  }

  plot(y, p[1,], type='l', lwd=2, xlab='Yield', ylab='Price', col=cl[1], ylim=c(min(p), max(p)))
  for(i in 2:4) {
  lines(y, p[i,], type='l', lwd=2, col=cl[i])
  }
  legend('topright', paste('cp = ', cp*100, sep=''), cex=.8, lwd=2, bty='n', col=cl)
```

```{r, label=fig-PY-Coupon, fig=T, echo=F}
  <<PY-Coupon>>
```

Figure \[fig:PY-Maturity\] represents the price-yield relationship for
10% bonds with maturities ranging from 3 to 30 years. Regardless of
maturity, the bonds are worth 100 when the yield is identical to the
coupon. This can be easily established by writing equation
(\[eq:fixedinBonds-1\]) as:

$$P(r) = N \left[\frac{c}{r} \left(1-\frac{1}{(1+r)^n}\right) + \frac{1}{(1+r)^n}  \right]$$

from which one concludes that $P(r) = N$ when $c=r$.

```{r, label=PY-Maturity, echo=T}
  ## plot price-yield as a function of maturity
  m <- c(3, 5, 10, 30)
  cl <- c('red', 'green', 'blue', 'black')
  y <- seq(0, .20, .01)
  cp <- .1
  p <- matrix(0, length(m), length(y))
  for(i in seq_along(m)) {
    p[i,] <- sapply(y, function(y) SimpleBondPrice(cp, m[i], yield=y))
  }

  plot(y, p[1,], type='l', lwd=2, xlab='Yield', ylab='Price', col=cl[1], ylim=c(min(p), max(p)))
  for(i in 2:4) {
  lines(y, p[i,], type='l', lwd=2, col=cl[i])
  }
  legend('topright', paste('m = ', m, sep=''), cex=.8, lwd=2, bty='n', col=cl)
```

```{r, fig-PY-Maturity, fig=T, echo=F}
  <<PY-Maturity>>
```

## Basic Measures of Interest Rate Risk

The interest rate risk of fixed income instruments is traditionally
measured by Variation, Sensitivity and Duration.

### Variation

Variation $V$ (also known as dollar duration) is the negative of the
derivative of price with respect to yield $$\begin{aligned}
V &=& -\frac{\partial P(r)}{\partial r} \\
&=& \sum_{i=1}^n \frac{i F_i}{(1+r)^{i+1}} \end{aligned}$$

```{r, }
  Variation <- function(coupon, n, N=100, yield) {
    100*sum(sapply(1:n, function(i) ifelse(i<n, i*coupon/(1 + yield)^(i+1), 
                                           n*(1+coupon)/(1+yield)^(n+1))))
  }
```

Variation is often measured by “PV01”, which is the change in bond price
for a 1 basis point change in yield:
$$\textrm{PV01} = V \times 10^{-4}$$

### Approximate Portfolio Yield

Variation is a useful tool for computing the approximate yield of a
portfolio, given the yield of each bond. Consider two bonds with prices
$P_1(r_1)$ and $P_2(r_2)$ and variations $V_1(r_1)$ and $V_2(r_2)$. We
look for the common yield $r^*$ such that:
$$P = a_1 P_1(r^*) + a_2 P_2(r^*)
\label{eq:fixedinBonds-2}$$

Use a first-order Taylor’s expansion of price $P_i(r)$ at $r_i$:
$$P_i(r^*) = P_i(r_i) + (r^*-r_i) V_i(r_i)$$

Replacing $P_i(r^*)$ by this approximation in (\[eq:fixedinBonds-2\])
yields:

$$\sum_{i=1}^2 a_i  (r^*-r_i) V_i(r_i) = 0$$

which gives
$$r^* = \frac{\sum_{i=1}^2 a_i  r_i V_i(r_i)}{\sum_{i=1}^2 a_i V_i(r_i)}$$

This approximation is valid when the yields $r_i$ are not too far apart,
as illustrated by the following calculation.

Consider a portfolio made of two bonds, described in
Table \[tab:fixedinBonds-1\].

  Bond   Coupon (%)   Maturity   Quantity   Yield (%)
  ------ ------------ ---------- ---------- -----------
  A      5            5          2          5.20
  B      6            7          3          5.60

The approximate portfolio yield is computed as follows:

```{r, label=portfolio-yield-1, echo=T}
  cp <- c(.05, .06); r <- c(.052, .056); T <- c(5,7); a<- c(2,3)
  P <- sapply(1:2, function(i) SimpleBondPrice(cp[i], T[i], yield=r[i]))
  V <- sapply(1:2, function(i) Variation(cp[i], T[i], yield=r[i]))

  r <- sum(r*V*a)/sum(V*a)
```

which gives an approximate yield of $r=`r round(r*100,2) ` \%$. An
exact calculation involves solving for $r^*$ the non-linear equation
$$P = \sum_{i=1}^2 a_i P_i(r^*)$$

```{r, label=portfolio-yield-2}
  fun <- function(r) {
    sum(P) - sum(sapply(1:2, function(i) SimpleBondPrice(cp[i], T[i], yield=r)))}

  r <- uniroot(fun, c(.05, .06))$root
```

which gives the exact portfolio yield $r= `r round(r*100,2) ` \%$.

### Sensitivity

The sensitivity $S$ (also called modified duration) is the percent
change in price per unit change in yield:
$$S = \frac{1}{P} V = \frac{1}{P(1+r)} \sum_{i=1}^n \frac{i F_i}{(1+r)^i}$$

```{r, }
  Sensitivity <- function(coupon, n, N=100, yield) {
    Variation(coupon,n,N,yield)/SimpleBondPrice(coupon, n, N, yield)
  }
```

### Modified Duration of a Portfolio

Consider a portfolio made of $n$ bonds, each with nominal amount $q_i$,
price $P_i$ and modified duration $S_i, i=1, \ldots, n$. Let $P$ be the
value of the portfolio, we want to determine the modified duration of
the portfolio

$$S = \frac{1}{P} \frac{\partial P}{\partial r}$$

$$\begin{aligned}
  S &=& \frac{1}{P} \sum_i q_i \frac{\partial P_i}{\partial r} \nonumber \\
  &=& \frac{1}{P} \sum_i q_i \frac{P_i}{P_i}\frac{\partial P_i}{\partial r} \nonumber \\
  &=& \frac{1}{P} \sum_i q_i P_i S_i \nonumber \\
  &=& \sum_i \frac{q_i P_i}{P}   S_i
\label{eq:mod-duration}\end{aligned}$$

The modified duration of a portfolio is the weighted average of the
modified durations of the bonds. The weight associated with each bond is
the fraction of portfolio value associated with the bond.

### Duration

The Macaulay duration $D$ (named after Frederick Macaulay who introduced
this risk measure), can be interpreted as the average maturity, weighted
by the present value of the cash flow at each payment date.

$$D = \sum_{i=1}^n \frac{i F_i}{P (1+r)^i}
\label{eq:macaulay}$$

The Macaulay Duration is closely related to Sensitivity:

$$S =  \frac{1}{(1+r)} D$$

```{r, }
  SimpleDuration <- function(coupon, n, yield) {
    100*sum(sapply(1:n, function(i) ifelse(i<n, i*coupon/(1 + yield)^i, 
                        n*(1+coupon)/(1+yield)^n)))/SimpleBondPrice(coupon,n,yield=yield)
  }
```

These measures can be used to estimate the approximate impact of yield
change on price. Using the previous example, the exact price change
caused by a $0.1\%$ yield increase is:

```{r, label=duration-1}
  d1 <- SimpleBondPrice(coupon=.06, n=10, yield=.051) - SimpleBondPrice(coupon=.06, n=10, yield=.05)
```

or $d_1 = `r round(d1,3) `$. An approximate calculation with Duration
yields:

```{r, label=duration-2}
  P <- SimpleBondPrice(coupon=.06, n=10, yield=.05)
  D <- SimpleDuration(coupon=.06, n=10, yield=.05)
  d2 <- -P * D/(1+.05) * .001
```

or a change in price of $d_2 = `r round(d2,3) `$ for a $.1\%$
increase in yield. Duration is a popular measure of price sensitivity
because of its intuitive interpretation: the price sensitivity of a bond
with duration $D$ is the same as the one of a zero-coupon bond maturing
in $D$ years.

It is a surprising fact that the duration of coupon bonds does not
increase indefinitely as maturity increases. The following script
computes the duration of bonds of varying maturities and coupon rates.

```{r, label=duration-maturity}
  coupon <- c(0, .1, .2, 2, 10, 20)/100
  maturity <- seq(from=1, to=100, by=1)
  yield <- .1
  res <- NULL
  for(c in coupon){
    res <- rbind(res, sapply(maturity, function(n) SimpleDuration(c, n, yield)))
    }
```

The resulting values are plotted in Figure  \[fig:dur-mat\]. As
expected, the duration of the zero-coupon bond is identical to its
maturity. For all other bonds, the duration converges to a constant as
maturity increases, regardless of coupon rate.

```{r, duration-maturity-plot, eval=FALSE, echo=FALSE}
  col <- array(c('red', 'green', 'blue', 'black'), dim=c(1,length(coupon)))

  plot(maturity, res[1,], type='l', ylim=c(0, max(res)), xlab='maturity', ylab='duration', lwd=2, col=col[1])
  abline(h=(1+yield)/yield, lwd=2, col='black')
  for(i in 2:length(coupon)) {
    lines(maturity, res[i,], lwd=2, col=col[i])
  }
  legend('topleft', paste('cp = ', coupon*100, sep=''), cex=.8, lwd=2, bty='n', col=col )
```

```{r, label=fig-duration-maturity, echo=FALSE, fig=T}
  <<duration-maturity-plot>>
```

This can be formally shown by first deriving an analytical formula for
Duration. Let $c$ be the coupon rate and $r$ the bond yield. The price
of a bond with nominal of 1 is:

$$\begin{aligned}
P &=& \frac{c}{r} (1-(1+r)^{-n}) + (1+r)^{-n} \\
  &=& \frac{1}{r} \left(c(1-(1+r)^{-n}) + (1+r)^{-n} \right) \\
  &=& \frac{1}{r} g(r)\end{aligned}$$

with $g(r) = c(1-(1+r)^{-n}) + (1+r)^{-n}$.

Now, $$\begin{aligned}
  \frac{\partial \log P}{\partial r} &=& -\frac{1}{r} + \frac{g'(r)}{g(r)} \\
  &=& \frac{1}{P} \frac{\partial P}{\partial r} \end{aligned}$$

But, $$\begin{aligned}
  D &=& -\frac{1+r}{P} \frac{\partial P}{\partial r} \\
  &=& (1+r) \left[ \frac{1}{r} - \frac{g'(r)}{g(r)} \right] \\
  &=& \frac{1+r}{r} - (1+r)\frac{g'(r)}{g(r)} \\
  &=& \frac{1+r}{r} - (1+r)\frac{n(c-r)(1+r)^{-1}+1}{c((1+r)^n-1)+r} \\
  &=& \frac{1+r}{r} - \frac{n(c-r)+1+r}{c((1+r)^n-1)+r}
  \end{aligned}$$

The last expression shows that duration converges towards
$\frac{1+r}{r}$ as maturity goes to infinity.

### Duration of a Portfolio {#sec:duration-of-a-portfolio}

Consider a portfolio made of $n$ bonds, each with nominal amount $q_i$,
price $P_i$ and duration $D_i, i=1, \ldots, n$. Let $P$ be the value of
the portfolio, what is its Duration? Since Duration is related to
sensitivity by the relation $$D = (1+r) S$$

multiplying both sides of equation \[eq:mod-duration\] by $(1+r)$
yields:

$$D =  \sum_i \frac{q_i P_i}{P} D_i
\label{eq:portfolio-duration}$$

## Bond Clean and Dirty Price

Bonds prices are quoted in “flat” or “clean” price, and not in net
present value. The clean price being the net present value less the
accrued interest. Therefore, the bond quoted price is not the amount
that the buyer disburses when she buys a bond. The reason for this
apparent complication is illustrated in Figure \[fig:cd-1\], which plots
the change over time of a bond clean price and present value, with yield
held constant. Quoting a bond by net present value would be impractical,
because of the discontinuity around the coupon payment date. The clean
price, however, remains steady over time, when yield is held constant.

The plots are obtained as follows, using the function of package :

```{r, label=CleanDirty, results='hide', warning=FALSE, message=FALSE}
  # define a bond
  b375 <- Bond('375', myDate('01jan2000'), myDate('04jan2019'), 0.0375, 100, 'a')
  b300 <- Bond('300', myDate('01jan2000'), myDate('04jan2019'), 0.0300, 100, 'a')
  dtToday <- myDate('25mar2010')
  dt <- timeSequence(dtToday, myDate('04jan2019'), by='day')
  dt <- dt[seq(1, length(dt),by=4)]

  # compute series of clean price, accrued interest and present value
  # for given yield
  bondcalc <- function(bond, y) {
  p <- sapply(dt, function(d) BondYield2Price(bond, d, y))
  ac <- rep(NA, length(p))
  for(i in seq(length(p))){ac[i] <- BondAC(bond, as.Date(dt[i]))}
  cp <- p-ac

  list(p=p, ac=ac, cp=cp)
  }
  bc1 <- bondcalc(b300, .03)
  bc2 <- bondcalc(b375, .03)
```

```{r, par-bond,echo=FALSE, fig.width=3, fig.height=3}
  plot(dt, bc1$p, type='l', lwd=2, xlab='', 
       ylab='Price', col='red', ylim=c(98, max(bc2$p)))
  lines(dt, bc1$cp, type='l', lwd=2, col='blue')
```

```{r, premium-bond, echo=FALSE,fig.width=3, fig.height=3}
  plot(dt, bc2$p, type='l', lwd=2, xlab='Settlement date', ylab='', 
  col='red', ylim=c(98, max(bc2$p)))
  lines(dt, bc2$cp, type='l', lwd=2, col='blue')
```
