---
editor_options: 
  markdown: 
    wrap: 72
---

# Bond Futures

```{r, IRFutures,echo=FALSE,results='hide'}
  library(empfin)
```

The Euro-Bund Futures contract is an agreement to buy at a fixed price,
at a future date, a "notional bond", with a 6% coupon at a 10 year
maturity.

## Forward Price

The forward price is a "clean price". Let:

$P_t^d$

:   Dirty price

$P_t^c$

:   Clean price

$AI_t$

:   Accrued interest

To compute the forward clean price at time $T$:

1.  No coupon payments between now and $T$:
    $$P_t^c = P_0^d e^{r_{0, T} T} - AI_T$$

2.  There is a coupon payment before $T$:

    1.  Compute $I$: present value of next coupon

    2.  Forward clean price:
        $$P_T^d = (P_0^d - I) e^{r_{0, T} T} - AI_T$$

<!-- -->

```{r, label=bond-1, echo=TRUE}
  b1 <- Bond(id='b1', dtIssue=myDate('01jan2000'), dtMaturity=myDate('15dec2018'),
             couponRate=.05, nominal=100, frequency='a')

  # price on coupon date
  dtSettlement <- myDate('17dec2010')
  p <- BondYield2Price(b1, dtSettlement, .05)
  ai <- BondAC(b1, dtSettlement)
  print(paste('DtSettlement:', dtSettlement, ' Dirty: ', round(p,2), ' AI: ' , round(ai,2),
              ' Clean: ' , round(p-ai,2)))

  # price in middle of coupon period
  dtSettlement <- myDate('17jun2010')
  p <- BondYield2Price(b1, dtSettlement, .05)
  ai <- BondAC(b1, dtSettlement)
  print(paste('DtSettlement:', dtSettlement, ' Dirty: ', round(p,2), ' AI: ' , round(ai,2),
              ' Clean: ' , round(p-ai,2)))
```

## Delivery against the Futures contract

A short position on a Futures contract is settled by delivering a
certain quantity of one of the eligible bonds in exchange for the
payment of the delivery price. The quantity is defined by a conversion
factor to account for differences in maturity and coupon.

Table $$tab:bund-1$$ shows the eligible bunds for delivery against the
June 2010 Euro-bund contract. Data is as of 25-mar-2010:

| Bund | Coupon | Maturity | Factor ($k_i$) | Yield |
|:-----|:-------|:---------|:---------------|------:|
| 374  | 3.75   | 4/1/2019 | 0.852328       |  3.00 |
| 382  | 3.50   | 4/7/2019 | 0.828936       |  3.05 |
| 390  | 3.25   | 4/1/2020 | 0.803710       |  3.10 |

The actual delivery cost, also called invoice amount is:

$$IP_T = P_T^c  \times CF + AI_T$$

where $CF$ is the conversion factor specific to each bond.

The mechanism is illustrated by the following example. On 25-Mar-2010,
the June Euro Bund contract trades at 123.24 on Eurex
(www.eurexchange.com). Delivery to take place on 10-Jun-2010.

For Bund 374, the accrued interest on 10-Jun-2010 is approximately:

```{r, Bund374}
  ai = 3.00 * as.numeric(myDate('10jun2010')-myDate('04Jan2010'))/365
  ai
```

On 25-Mar-2010, you sell one Euro-Bund Futures at 123.24. On delivery
day, 10-Jun-2010, you decide to use Bund 374 to settle your short
position. You receive the delivery price:

```{r, DeliveryPrice}
  Ft <- 123.24
  k <- .852328
  DelPrice <- Ft * k + ai
  DelPrice
```

We can compute the cost of delivering Bund 374 under various yield
scenarios on 10Jun2010.

## The Conversion Factor

The conversion factor is determined so that, for a flat yield curve at
$6\%$ and a Futures trading at par (100), all bonds are equivalent, as
far as delivery cost is concerned. This can be easily verified: Table []
compares the delivery price to the value of the delivered security for
each eligible bond.

## Determination of the Cheapest to Deliver

When the yield curve is not flat at the notional rate ($6\%$ for the
Euro-Bund), all bonds are no longer equivalent with respect to delivery
cost, and the seller will obviously determine the cheapest to deliver.
This is simply done by computing the cost associated with using each
bond for delivery.

With data from 25Mar2010, the results are as follows:

$$Insert table$$

which shows that delivering bund 374 is the cheapest way to settle a
short position.

It can be shown that when yields are below the notional yield, the
cheapest to deliver is the bond with the lowest duration. The inverse
holds true when yields are above the par yield.

## Implied Repo Rate

An alternate and more accurate method for determining the cheapest to
deliver is to compute the implied repo rate for each bond. The CTD bond
is the one that maximizes the total return of an arbitrageur that sells
the Futures, buys and hold a deliverable bond, and uses that bond to
settle his short Futures position. The bond with the highest IRR is the
cheapest to deliver.

Let:

$P^d_t$

:   Purchase price of bond at time $t$

$F_t$

:   Futures price

$c$

:   Conversion factor

$ai_t$

:   Accrued interest at time $t$

$P^i_T$

:   Invoice price, i.e. $F_t fc + ai_T$

The implied repo rate is the return from holding the bond to expiry of
the futures contract. The price of the Futures is linked to the IRR by
the following relation:

$$F_t c + ai_T =  P^d_t \left( 1+ IRR \frac{n_1}{n_2} \right)
    \label{eq:irr}$$

Table [] shows the calculation for each deliverable bond, and confirms
that Bund 374 is the cheapest to deliver.

## The Delivery Option

The seller has the option to choose the cheapest to deliver. The
approximate theoretical value of the futures is

$$F_t = \frac{P^C_T}{fc}$$

Figure [] shows the theoretical price of the Futures for 2 deliverable
bonds, as a function of the yield at settlement. The Futures price must
ensure that all deliverable bonds have a positive basis (otherwise,
there would be a riskless arbitrage). We see from figure [] that this
requirement creates a pattern in the choice of CTD. When yield is above
the notional coupon, the long maturity bond is the CTD. The inverse
holds true when yield is below the notional coupon.

## IR sensitivity of Bund Futures

From equation $$eq:irr$$, one can observe that the Futures price $F_t$
is a function of the short rate (). To correctly measure the IR
sensitivity of a IR Futures, one needs to consider the replicating
portfolio of a cash-carry trade:

1.  Today:

    1.  Borrow at the IRR from now until settlement of the Futures

    2.  Buy the CTD

2.  At delivery:

    1.  Deliver the CTD

    2.  Receive the Invoice amount

    3.  Pay the short term loan

## Cash-Carry Arbitrage

1.  Today:

    1.  Borrow at the IRR from now until settlement of the Futures

    2.  Buy the CTD

2.  At delivery:

    1.  Deliver the CTD

    2.  Receive the Invoice amount

    3.  Pay the short term loan

This identifies two sensitivities:

1.  Sensitivity to yield of CTD
    $$\frac{\partial F}{\partial y} = \frac{1}{c} \frac{\partial P^d_t}{\partial y} \left( 1+ IRR \frac{n_1}{n_2} \right)$$

2.  Sensitivity to IRR from today to settlement.
    $$\frac{\partial F}{\partial IRR} = \frac{P^d_t}{c} \frac{n_1}{n_2}$$

## Hedging a Bond with a Euro-Bund Contract

How to construct a hedge for a bond, using Euro-Bund contracts and
EURIBOR contracts is best explained through an example:

You want to hedge 100 million of notional amount of a 15-year to
maturity Bund ($B_1$) with the June Euro-Bund contract. You first
compute the PV01 of your asset and get:

Next, you determine the sensitivity of the Bund futures, assuming that
Bund 374 will be the CTD, and get:

1.  PV01 / bond yield:

2.  PV01 / IRR:

To match the PV01 with respect to bond yield, you must sell:

rounded to $x$ Euro-bund June 2010 futures. This creates an exposure to
the 3-month rate (IRR from 25mar2010 to 10Jun2010), which is hedged by y
million of q x w FRA.
