# The Model of Hull & White

```{r, init-HullWhite,echo=TRUE,results='hide'}
  library(fBasics)
```

Hull & White have developed a trinomial model for the short rate, that
can be exactly calibrated to the zero-coupon curve. The short rate
process is an Ornstein-Uhlenbeck (OU) mean reverting process, and we
start this section by a short review of the properties of this process.

## Moments of $x_t$ under a OU Process

To construct a Hull & White model of the short rate, we need to compute
the moments of a OU process:

$$dx_t = \theta(\mu-x_t) dt + \sigma dW_T$$

To compute $E(x_t)$, consider the function

$$f(x, t) = x e^{\theta t}$$

Apply Ito’s Lemma:

$$\begin{aligned}
df &=& (\theta x_t e^{\theta t} + \theta(\mu-x_t) e^{\theta t}) dt + \sigma e^{\theta t} dW \\
  &=& \theta \mu e^{\theta t} dt + \sigma e^{\theta t} dW
\end{aligned}$$

Integrate:

$$\begin{aligned}
x_t e^{\theta t} &=& x_0 + \theta \mu \int_0^t e^{\theta u} du + \sigma \int_0^t e^{\theta u} dW_u \\
&=& x_0 + \mu (e^{\theta t} - 1) +\sigma \int_0^t e^{\theta u} dW_u \\
x_t &=& x_0 e^{-\theta t} + \mu (1-e^{-\theta t}) +\sigma \int_0^t e^{-\theta (t-u)} dW_u
\end{aligned}$$

Thus,

$$E(x_t) = x_0 e^{-\theta t} + \mu (1-e^{-\theta t})
\label{eq:mean-ou}$$

The variance of $x_t$ is now easily computed:

$$\begin{aligned}
V(x_t) &=& E\left[ (x_t - E(x_t))^2 \right] \\
&=& E \left[ \sigma^2 \left( \int_0^t e^{-\theta(t-u)} dW_u \right)^2 \right] \\
&=& \sigma^2 e^{-2\theta t} E \left[  \left( \int_0^t e^{\theta u} dW_u \right)^2 \right] \\
&=& \sigma^2 e^{-2\theta t} \int_0^t e^{2\theta u} du
\end{aligned}$$

Finally,

$$V(x_t) = \frac{\sigma^2}{2 \theta} (1-e^{-2 \theta t})
\label{eq:var-ou}$$

## The Hull & White Trinomial Model

The Hull & White model defines an OU process for the short rate:

$$d r_t = (\theta(t) - a r_t) dt + \sigma dz_t$$

This is an generalization of Vasicek’s model, which defined the
following process for the short rate:

$$d r_t = a(b - r_t) dt + \sigma dz_t$$

Using results from Vasicek’s, one could compute an analytical expression
for $\theta(t)$ such that the process is consistent with observed
zero-coupon yields. Rather than doing that, however, we will construct a
discretized representation of the process with a trinomial tree, and
calibrate the parameters of that tree so that the discretized model
reprices exactly the zero-coupons.

The tree is constructed in two steps:

1.  A trinomial tree is constructed for the process
    $$d r^*_t = -a r^*_t dt + \sigma dz_t$$

2.  At each time step, the process is shifted to reprice the zero-coupon
    of that maturity: $$r_t = r^*_t + \alpha_t$$

Each step is now described.

### Step 1

The trinomial tree has time increments $\Delta t$. Nodes are evenly
spaced with an increment $\Delta r = \sigma \sqrt{3 \Delta t}$. Node
$(i,j)$ corresponds to time step $i \Delta t$ and rate level
$j \Delta r$.

From \[eq:mean-ou\] and \[eq:var-ou\],

$$\begin{aligned}
  E(\Delta r^*) &=& r^*_t (e^{-a \Delta t}-1) \nonumber \\
  &=& -a r^* \Delta t \nonumber \\
  V(\Delta r^*) &=& \frac{\sigma^2}{2a} (1-e^{-2a \Delta t}) \nonumber \\
  &=& \sigma^2 \Delta t
\end{aligned}$$

Starting at node $(i,j)$, with rate $r^*_t = j \Delta r$, the rate can
move to three values at step
$(i+1): r^*_t + \Delta r, r^*_t, r^*_t - \Delta r$. We now compute the
transition probabilities by matching the first two moments.

$$\begin{aligned}
  E(\Delta r^*) &=& p_u \Delta r - p_d \Delta r \\
&=& -ar^* \Delta t \\
&=& -a j \Delta r \Delta t \\
V(\Delta r^*) &=& p_u \Delta r^2 + p_d \Delta r^2 - E(\Delta r)^2 \\
&=& \sigma^2 \Delta t
\end{aligned}$$

After some algebra, one obtains the following system of linear
equations:

$$\begin{aligned}
    p_u + p_d &=& \frac{1}{3} + a^2 j^2 \Delta t^2 \\
    p_u + p_d &=& \frac{1}{3} + a^2 j^2 \Delta t^2-aj \Delta t
  \end{aligned}$$

Finally,

$$\begin{aligned}
    p_u &=& \frac{1}{6} + \frac{a^2 j^2 \Delta t^2 - aj \Delta t}{2} \\
    p_u &=& \frac{1}{6} + \frac{a^2 j^2 \Delta t^2 + aj \Delta t}{2} \\
    p_m &=& 1-p_u - p_d
  \end{aligned}$$

This calculation is illustrated below:

```{r, label=HW-1, echo=T}
  a <- .1; dt <- 1; s <- .01

  pu <- function(j) {
    (1/6) + (a^2 * j^2 * dt^2 - a*j*dt)/2}

  pd <- function(j) {
    (1/6) + (a^2 * j^2 * dt^2 + a*j*dt)/2}

  pm <- function(j) {
    1-pu(j) - pd(j)}


  j <- 0
  print(paste('pu: ', pu(j), ' pm: ', pm(j), ' pd: ', pd(j)))
```

Sadly, the same calculation at node $(9,-9)$ yields unfortunate results:

```{r, label=HW-2, echo=T}
  j <- -9
  print(paste('pu: ', pu(j), ' pm: ', pm(j), ' pd: ', pd(j)))
```

At this node, we need to change the branching pattern to ensure positive
probabilities. We compute now the transition probabilities when the
future states of $r^*$ are $r^*+2\Delta r, r^*+ \Delta r, r^*$. The
moment-matching equations become:

$$\begin{aligned}
  E(\Delta r^*) &=& p_u 2\Delta r + p_m \Delta r \\
&=& -a j \Delta r \Delta t \\
V(\Delta r^*) &=& p_u (2 \Delta )r^2 + p_m \Delta r^2 - E(\Delta r)^2 \\
&=& \sigma^2 \Delta t
\end{aligned}$$

After some algebra, one gets:

$$\begin{aligned}
    p_u &=& \frac{1}{6} + \frac{a^2 j^2 \Delta t^2 + aj \Delta t}{2} \\
    p_m &=& 1 - p_u - p_d \\
    p_d &=& \frac{7}{6} + \frac{a^2 j^2 \Delta t^2 - 3aj \Delta t}{2}
  \end{aligned}$$

And we verify that the probabilities are now good:

```{r, label=HW-3, echo=T}
  pu2 <- function(j) {
    (1/6) + (a^2 * j^2 * dt^2 + a*j*dt)/2}

  pd2 <- function(j) {
    (7/6) + (a^2 * j^2 * dt^2 + 3*a*j*dt)/2}

  pm2 <- function(j) {
    1-pu2(j) - pd2(j)}

  j <- -9
  print(paste('pu: ', pu2(j), ' pm: ', pm2(j), ' pd: ', pd2(j)))
```

As a rule, the geometry at each node can be adapted to ensure positive
probabilities.

### Step 2

In the next step, we transform the $r^*$ process into the actual process
for the short rate by adding a constant $\alpha_i$ to all nodes at time
step $i\Delta t$. The calculation of constants $\alpha_i$ is performed
recursively, one time step at a time. We assume that zero-coupon rates
are known for each maturity $i \Delta t$. The corresponding zero-coupon
prices are:

$$P(0, i \Delta t) = e^{-R(0, i \Delta t) i \Delta t}$$

The short rate at the root of the tree, $r_{0,0}$ is by definition
$R(0, \Delta t)$, therefore:

$$\alpha_0 = R(0, \Delta t)$$

Let $Q_{i,j}$ be the price at time 0 of the Arrow-Debreu security for
state $(i,j)$. We now write the price of a zero-coupon maturing at
$2 \Delta t$.

$$P(0,2 \Delta t) = Q_{1,1} e^{-(\alpha_1 +\Delta r) \Delta t} +
Q_{1,0} e^{-\alpha_1 \Delta t} +
Q_{1,-1} e^{-(\alpha_1 -\Delta r) \Delta t}$$

which yields:

$$\alpha_1 = \frac{1}{\Delta t} \ln \left[ \frac{Q_{1,1} e^{-\Delta r \Delta t} +
Q_{1,0} +
Q_{1,-1} e^{\Delta r \Delta t}}{P(0, 2 \Delta t)} \right]$$

The generic iteration at step $i$ is:

1.  Solve for $\alpha_i$:
    $$P(0, (m+1) \Delta t) = \sum_{j=-m}^m Q_{m,j} e^{-(\alpha_m + j \Delta r) \Delta t}$$

2.  Compute the state prices for time step $m$:
    $$Q_{m+1, j} = \sum_k Q_{m,k} p_{k,j} e^{-(\alpha_m + k \Delta r) \Delta t}$$


## Generalization

The Hull & White model can be generalized to richer short rate
processes. For any short rate process of the form

$$df(r_t) = (\theta(t) - a f(r_t)) dt + \sigma dz_t$$

where $f()$ is invertible, define $x = f(r)$ to get:

$$dx_t = (\theta(t) - a x_t) dt + \sigma dz_t$$

Stage 1 above can be applied to the process $dx_t$. Stage 2 is slightly
modified: let $g(x) = f^{-1}(x)$. The equation to be solved for
$\alpha_i$ is now:

$$P(0, (m+1) \Delta t) = \sum_{j=-m}^m Q_{m,j} e^{-g(\alpha_m + j \Delta x) \Delta t}$$

Setting $f(x) = \ln(x)$ is particularly interesting, because in does not
allow negative interest rates.
