---
title: "Untitled"
author: "P. Henaff"
date: "2023-03-09"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo=TRUE}
dist <- function(x) sqrt(sum(x^2))
R <- 1
volume.sphere <- function(n) pi^(n/2) * R^n / gamma(n/2 + 1)
volume.cube <- function(n) (2*R)^n

dim <- 5
vol.sphere.e <- volume.sphere(dim)
vol.cube.e <- volume.cube(dim)
N <- 100000
in.sphere <- 0
for(i in seq(N)) {
  x <- runif(dim,0,1)
  in.sphere = in.sphere + ifelse(dist(x)<=1,1,0)
}

vol.sphere.s <- vol.cube.e * in.sphere / N
print(paste("exact:", vol.sphere.e, "sim:", vol.sphere.s))
```

```{r}
k <- round(N^(1/dim))
int <- 2/k
tmp <- rep(list(seq(-1+int/2,1-int/2,by=int)), dim)
grid <- expand.grid(tmp)
for(i in 1:dim(grid)[1]) {
  in.sphere = in.sphere + ifelse(dist(grid[i,])<=1,1,0)
}

vol.sphere.s <- vol.cube.e * in.sphere / N
print(paste("exact:", vol.sphere.e, "sim:", vol.sphere.s))
```


```

```{r, echo=FALSE, cache=TRUE}
registerDoParallel(used.cores)

calc.r <- function(N, algo="sobol") {
  nb.trials <- N/batch.size
  stats.in <- foreach(icount(nb.trials), .combine=rbind) %dopar% {
    if(algo == "sobol") {
      x <- sobol(batch.size, dim=dim, init=FALSE)
    } else {
      x <- runif(batch.size*dim, 0, 1)
      x <- matrix(x, nrow=batch.size)
    }
    nb.inside(x)
}
vol.cube.e * sum(stats.in[,1]) / N
}

R <- 1

dim <- 3
vol.sphere.e <- volume.sphere(dim)
vol.cube.e <- volume.cube(dim)

n.expo <- seq(14,20)
nb.samples <- length(n.expo)
batch.size <- 10000
res.d <- numeric(nb.samples)
res.s <- numeric(nb.samples)
for(i in seq(nb.samples)) {
  N <- 2^n.expo[i] - 1
  res.d[i] <- calc.r(N, "default")
  res.s[i] <- calc.r(N, "sobol")
}

N.seq <- 2^n.expo - 1
```

```{r rand-sobol, echo=FALSE, warning=FALSE} 
plot(N.seq, res.d, type="l", col="red", lwd=2,ylim=c(.95*vol.sphere.e, 1.05*vol.sphere.e), xlab="Number of points", ylab="Estimated volume of sphere")
lines(N.seq, res.s, type="l", col="green", lwd=2)
abline(h=vol.sphere.e, col="black", type="-", lwd=1)
legend("topright", legend=c("Default", "Sobol"),
       col=c("red", "green"), lty=c(1,1), cex=0.8)
```


  
}
```
